---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 6
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
library(tidyr)
```

---

## 1. Market Narrative (NASDAQ 2022-2023)

In financial markets, structural shifts often reflect the transition between macro cycles. By including data from late 2022, we can model the transition from the bear market into the AI-driven recovery of 2023.

### The Data: Oct 2022 to Dec 2023
We use monthly closing prices for Apple, Microsoft, and Nvidia.
```{r real-data}
dat_market <- data.frame(
  month  = rep(1:15, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 15),
  price  = c(
    # NVDA
    13.50, 16.50, 14.60, 19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    # MSFT
    232, 255, 240, 241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    # AAPL
    153, 148, 130, 142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)
dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))
```

### HMC Discovery with `smoothbp()`
To ensure we capture the sharpest price movements, we use **five breakpoints** and the full HMC engine without spike-and-slab regularization.

```{r fit-market-hmc}
fit_market <- smoothbp(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker), 
  omega  = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker),
  rho    = list(~ 1, ~ 1, ~ 1, ~ 1, ~ 1), 
  data   = dat_market,
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50), 
    delta = prior_normal(0, 80), 
    omega = list(
      prior_normal(2, 1, lb=1, ub=15), prior_normal(5, 1, lb=1, ub=15),
      prior_normal(8, 1, lb=1, ub=15), prior_normal(12, 1, lb=1, ub=15),
      prior_normal(14, 1, lb=1, ub=15)
    ),
    rho = list(prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1))
  ),
  chains = 4, iter = 5000, warmup = 3000
)

# Extract Ticker-Specific Breakpoints
draws <- as_draws_matrix(fit_market$draws)
om_cols <- colnames(draws)[grepl("^omega", colnames(draws))]
om_summary <- summarise_draws(draws[, om_cols]) |>
  separate(variable, into = c("bp", "param"), sep = "_") |>
  mutate(
    event = case_when(
      bp == "omega1" ~ "Bear Pivot",
      bp == "omega2" ~ "March Dip",
      bp == "omega3" ~ "AI Pivot",
      bp == "omega4" ~ "Autumn Dip",
      bp == "omega5" ~ "Year-End",
      TRUE ~ bp
    ),
    ticker = gsub("ticker", "", param),
    ticker = ifelse(ticker == "(Intercept)", "AAPL", ticker)
  )

# Calculate the actual omega for each ticker
# (Base omega + offset)
om_pivoted <- om_summary |>
  select(event, ticker, mean) |>
  pivot_wider(names_from = ticker, values_from = mean) |>
  mutate(
    MSFT = AAPL + MSFT,
    NVDA = AAPL + NVDA
  ) |>
  pivot_longer(cols = -event, names_to = "ticker", values_to = "omega_val")

# Merge with fitted values
pred <- fitted(fit_market)
dat_market$y_fit <- pred$fitted_mean
dat_market$lo    <- pred$fitted_Q2.5
dat_market$hi    <- pred$fitted_Q97.5

# Calculate y-position for the X markers (on the fitted line)
dat_markers <- om_pivoted |>
  rowwise() |>
  mutate(y_val = approx(dat_market$month[dat_market$ticker == ticker], 
                        dat_market$y_fit[dat_market$ticker == ticker], 
                        xout = omega_val)$y)

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point(alpha = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  # Red X for each breakpoint
  geom_point(data = dat_markers, aes(x = omega_val, y = y_val), color = "red", shape = 4, size = 4, stroke = 1.5) +
  scale_x_continuous(breaks = 1:15, labels = c("Oct22", "Nov", "Dec", "Jan23", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "High-Fidelity Market Narrative", subtitle = "Red 'X' marks the structural change-points discovered for each security.")
```

### Quantifying Market Leadership
Using the discovered pivot dates for Event 3 (The AI Pivot), we can see how the lead/lag relationship evolved.

```{r market-prob}
p_nvda_lead <- mean((draws[, "omega3_tickerNVDA"] - draws[, "omega3_tickerMSFT"]) < 0)
message(sprintf("Probability NVDA led the AI Pivot (Event 3): %.2f", p_nvda_lead))
```

---

## 2. Shifting Change-Points (Non-stationary Locations)

`smoothbp` also allows the change-point location ($\omega$) to depend on covariates.

```{r fit-shifting}
set.seed(456)
n_subj <- 40; n_obs <- 10; tau <- seq(0, 10, length.out = n_obs); x <- runif(n_subj, 0, 1)
rows <- list()
for (i in 1:n_subj) {
  om_i <- 7 - 4 * x[i]; di <- tau - om_i
  yi   <- 5 + 0.3 * di - 1.2 * di * plogis(di * 5) + rnorm(n_obs, sd = 0.1)
  rows[[i]] <- data.frame(subject = i, tau = tau, y = yi, x = x[i])
}
dat_shift <- do.call(rbind, rows)

fit_shift <- smoothbp(
  formula = y ~ tau, b0 = ~ 1 + (1 | subject), omega = list(~ x), deltas = list(~ 1),
  data = dat_shift, priors = smoothbp_priors(omega = list(prior_normal(5, 3, lb = 0))),
  chains = 2, iter = 2000, warmup = 1000, seed = 42
)
summary(fit_shift, effects = "fixed") |> filter(grepl("^omega", variable))
```

---

## Conclusion

By treating $\delta$, $\omega$, and $\rho$ as parameters that can vary with predictors, `smoothbp` provides a unified framework for modeling structural changes in complex systems.
