---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 6
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
library(tidyr)
```

---

## 1. Market Turning Points (NASDAQ 2022-2023)

In this example, we model the structural evolution of three tech giants. Rather than simply comparing the "center" of transitions ($\omega$), we derive the **actual turning points**—the local maxima and minima where the price trend reverses.

### The Data
Monthly closing prices from Oct 2022 to Dec 2023.
```{r real-data}
dat_market <- data.frame(
  month  = rep(1:15, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 15),
  price  = c(
    13.50, 16.50, 14.60, 19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    232, 255, 240, 241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    153, 148, 130, 142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)
dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))
```

### Fitting the Model
We use `smoothbp()` with five candidate breakpoints to capture the high-frequency structural shifts of 2023.

```{r fit-market}
fit_market <- smoothbp(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker), 
  omega  = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker),
  rho    = list(~ 1, ~ 1, ~ 1, ~ 1, ~ 1), 
  data   = dat_market,
    priors = smoothbp_priors(
      b0 = prior_normal(200, 500), b1 = prior_normal(0, 50), deltas = prior_normal(0, 80),
      # Constrain omegas: intercepts centered on windows, narrow priors for ticker offsets
      omega = lapply(list(3, 6, 9, 12, 14), function(m) {
        list("(Intercept)" = prior_normal(m, 1, lb = 1, ub = 15),
             "tickerMSFT"  = prior_normal(0, 1),
             "tickerNVDA"  = prior_normal(0, 1))
      }),
      rho = prior_normal(20, 5, lb = 5)
    ),
  chains = 4, iter = 4000, warmup = 2000
)
```

### The Point of Maximum Structural Curvature

The `smoothbp` mean function is:
$$\mu(\tau) = b_0 + b_1(\tau - \omega_1) + \sum_{k=1}^K \delta_k (\tau - \omega_k)\, \sigma(\rho_k(\tau - \omega_k))$$

The $k$-th breakpoint contributes a smooth, S-shaped transition in slope centred on $\omega_k$. The second derivative $f''(\tau)$ — measuring the *rate of bending* — has its maximum **exactly at $\tau = \omega_k$** by the symmetry of the logistic function. This holds regardless of the sign of $\delta_k$: a breakpoint that accelerates a rally and one that decelerates it both have their maximum curvature at $\omega_k$.

This means **$\omega_k$ directly answers "when did the structural change happen most rapidly?"** — no root-finding required.

### Extracting Omega per Ticker

Because `omega` is modelled as `~ ticker`, each ticker has its own posterior over $\omega_k$, representing when *its* structural transition was fastest.

```{r event-omegas}
n_bp    <- 5
draws   <- as_draws_matrix(fit_market$draws)
tickers <- levels(dat_market$ticker)

# Helper to safely extract a named draw column (handles special chars via grep)
val <- function(d, pat) {
  v <- d[grep(paste0("^", pat, "$"), names(d))]
  if (length(v) == 0) return(0)
  unname(v[1])
}

# Extract omega_k for each ticker from a single draw row
get_omegas <- function(d, tk) {
  sapply(1:n_bp, function(k) {
    base <- val(d, paste0("omega", k, "_\\(Intercept\\)"))
    off  <- if (tk == tickers[1]) 0 else val(d, paste0("omega", k, "_ticker", tk))
    base + off
  })
}

# Extract all ticker-specific parameters for a single draw row
get_params <- function(d, tk) {
  b1 <- val(d, "b1_\\(Intercept\\)")
  if (tk != tickers[1]) b1 <- b1 + val(d, paste0("b1_ticker", tk))
  
  deltas <- omegas <- rhos <- numeric(n_bp)
  for (k in 1:n_bp) {
    deltas[k] <- val(d, paste0("delta", k, "_\\(Intercept\\)"))
    omegas[k] <- val(d, paste0("omega", k, "_\\(Intercept\\)"))
    rhos[k]   <- val(d, paste0("rho", k, "_\\(Intercept\\)"))
    
    if (tk != tickers[1]) {
      deltas[k] <- deltas[k] + val(d, paste0("delta", k, "_ticker", tk))
      omegas[k] <- omegas[k] + val(d, paste0("omega", k, "_ticker", tk))
      rhos[k]   <- rhos[k]   + val(d, paste0("rho", k, "_ticker", tk))
    }
  }
  list(b1 = b1, deltas = deltas, omegas = omegas, rhos = rhos)
}

# Posterior summary of omega_k per ticker
omega_summary <- do.call(rbind, lapply(tickers, function(tk) {
  om_mat <- t(apply(draws, 1, get_omegas, tk = tk))
  do.call(rbind, lapply(1:n_bp, function(k) {
    data.frame(
      ticker = tk, bp = k,
      mean  = mean(om_mat[, k]),
      Q2.5  = quantile(om_mat[, k], 0.025),
      Q97.5 = quantile(om_mat[, k], 0.975)
    )
  }))
})) |> mutate(ticker = factor(ticker, levels = levels(dat_market$ticker)))
print(omega_summary)
```

### Visualising Structural Events

We mark each $\omega_k$ on the fitted curve — the point of maximum curvature — for every ticker. Horizontal bars show the 95% credible interval.

```{r plot-events}
pred_orig        <- fitted(fit_market)
dat_market$y_fit <- pred_orig$fitted_mean
dat_market$lo    <- pred_orig$fitted_Q2.5
dat_market$hi    <- pred_orig$fitted_Q97.5

# Compute y-position on the mean curve at each omega
omega_plot <- omega_summary |>
  rowwise() |>
  mutate(y_val = approx(
    dat_market$month[dat_market$ticker == ticker],
    dat_market$y_fit[dat_market$ticker  == ticker],
    xout = mean)$y) |>
  ungroup()

month_labels <- c("Oct22","Nov","Dec","Jan23","Feb","Mar","Apr","May",
                  "Jun","Jul","Aug","Sep","Oct","Nov","Dec")

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point(alpha = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  geom_point(
    data = omega_plot, aes(x = mean, y = y_val),
    color = "red", shape = 4, size = 4, stroke = 1.5, inherit.aes = FALSE
  ) +
  geom_errorbarh(
    data = omega_plot, aes(xmin = Q2.5, xmax = Q97.5, y = y_val),
    color = "red", height = 0, alpha = 0.4, inherit.aes = FALSE
  ) +
  scale_x_continuous(breaks = 1:15, labels = month_labels) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Structural Events: Point of Maximum Curvature (omega_k)",
    subtitle = "Red X = posterior mean omega_k; bars = 95% CI."
  )
```

### Targeted Leadership Analysis

Having identified the structural events visually, we can now perform targeted hypothesis testing on the posterior distribution. We focus on three key questions:
1.  **The Early Pivot**: Who turned first in late 2022 (BP 1)?
2.  **The Autumn Shift**: Who reacted first to the late-2023 market dip (BP 4)?
3.  **Cross-Event Dynamics**: Did NVIDIA's primary surge (BP 2) happen before Apple's initial recovery (BP 1)?

```{r leadership-analysis}
# Extract omegas for all draws for the targeted indices
om_nvda <- t(apply(draws, 1, get_omegas, tk = "NVDA"))
om_msft <- t(apply(draws, 1, get_omegas, tk = "MSFT"))
om_aapl <- t(apply(draws, 1, get_omegas, tk = "AAPL"))

# 1. Early Pivot (BP 1)
p1_nvda_msft <- mean(om_nvda[, 1] < om_msft[, 1])
p1_nvda_aapl <- mean(om_nvda[, 1] < om_aapl[, 1])

# 2. Autumn Shift (BP 4)
p4_nvda_msft <- mean(om_nvda[, 4] < om_msft[, 4])
p4_msft_aapl <- mean(om_msft[, 4] < om_aapl[, 4])

# 3. Cross-Event: NVDA Surge (BP 2) vs AAPL Recovery (BP 1)
p_cross <- mean(om_nvda[, 2] < om_aapl[, 1])

cat("--- Event 1: Early Pivot (BP 1) ---\n")
message(sprintf("Prob NVDA led MSFT: %.2f", p1_nvda_msft))
message(sprintf("Prob NVDA led AAPL: %.2f", p1_nvda_aapl))

cat("\n--- Event 2: Autumn Shift (BP 4) ---\n")
message(sprintf("Prob NVDA led MSFT: %.2f", p4_nvda_msft))
message(sprintf("Prob MSFT led AAPL: %.2f", p4_msft_aapl))

cat("\n--- Cross-Event Dynamics ---\n")
message(sprintf("Prob NVDA AI Surge (BP 2) led AAPL Recovery (BP 1): %.2f", p_cross))
```


Note that these probabilities are over the *timing of maximum structural curvature* within the calendar window — not whether any slope ever changed sign. If the model finds NVIDIA's transition was fastest in February while Microsoft's was fastest in March, we conclude NVIDIA led the rally.

---

## Conclusion

By recognising that $\omega_k$ is analytically the point of maximum structural curvature, `smoothbp` gives us a principled, closed-form basis for lead/lag analysis. The key challenge in multi-ticker models is not estimation but *identification*: ensuring we compare the same calendar event across tickers. Temporal-window matching, informed by domain knowledge of when events occurred, is the robust solution to this identification problem.

