---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 4
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
```

---

## 1. Conditional Breakpoints (Emergent Behavior)

A common challenge is when the **existence** of a breakpoint depends on a covariate or experimental group. Instead of fitting separate models, we can use interaction terms in the `deltas` sub-model paired with spike-and-slab priors.

### Simulation: Group-specific structure
* **Group A (Control)**: Simple linear trend (0 BP).
* **Group B (Treatment)**: One smooth breakpoint at $\omega=5$.

```{r sim-conditional}
set.seed(123)
n <- 150
tau <- seq(0, 10, length.out = n)

# Group A: Linear
yA <- 2 + 0.5 * tau + rnorm(n, sd = 0.2)

# Group B: 1 BP at 5
om <- 5; rho <- 4; delta <- -1.5
di <- tau - om; si <- plogis(di * rho)
yB <- 2 + 0.5 * di + delta * di * si + rnorm(n, sd = 0.2)

dat_cond <- data.frame(
  y     = c(yA, yB),
  tau   = rep(tau, 2),
  group = rep(c("A", "B"), each = n)
)
```

### Fitting with Spike-and-Slab Interactions
By using `deltas = list(~ group)`, the model can learn different slope changes for each group.

```{r fit-conditional}
fit_cond <- smoothbp_ss(
  formula = y ~ tau,
  b0      = ~ group,
  b1      = ~ group,
  deltas  = list(~ group),
  omega   = list(~ 1),
  rho     = list(~ 1),
  data    = dat_cond,
  spike   = prior_spike_slab(pi = 0.2, learn_pi = TRUE),
  chains  = 2, iter = 2000, warmup = 1000, seed = 42
)

# Visualize observed vs fitted
pred_cond <- fitted(fit_cond)
dat_cond$y_fit <- pred_cond$fitted_mean
dat_cond$lo    <- pred_cond$fitted_Q2.5
dat_cond$hi    <- pred_cond$fitted_Q97.5

ggplot(dat_cond, aes(x = tau, y = y, color = group)) +
  geom_point(alpha = 0.2) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = group), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  facet_wrap(~group) +
  theme_minimal() +
  labs(title = "Conditional Breakpoint Selection", subtitle = "Control correctly identified as linear; Treatment captures the break.")
```

---

## 2. Market Leader-Follower Dynamics (Real NASDAQ 2023 Data)

In financial time series, assets often react to macro trends with varying lags. We use **real 2023 monthly closing prices** for Nvidia (NVDA), Microsoft (MSFT), and Apple (AAPL). 

The 2023 market was characterized by two major shifts:
1.  **The AI Pivot**: A structural trend shift in late spring/early summer.
2.  **The Year-End Rally**: A sharp recovery starting in late autumn.

### The Data: AI Boom and Year-End Recovery
We use the actual monthly closing prices to model these two distinct shifts.

```{r real-data}
dat_market <- data.frame(
  month  = rep(1:12, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 12),
  price  = c(
    # NVDA
    19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    # MSFT
    241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    # AAPL
    142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point() + geom_line() +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() +
  labs(title = "Real 2023 Monthly Closing Prices", y = "Price (USD)", x = "Month")
```

### Modeling Philosophy: Noise vs. Structure
We can fit the model in two ways: one that treats mid-year volatility as noise (conservative prior) and one that treats it as a structural event to be captured (permissive prior).

#### Option 1: The "Conservative" Model (Smooth Trend)
Here we use a low inclusion probability (`pi = 0.1`), forcing the model to only "accept" a second breakpoint if the evidence is overwhelming.

```{r fit-market-cons}
dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))

fit_cons <- smoothbp_ss(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker), omega = list(~ ticker, ~ ticker),
  rho = list(~ 1, ~ 1), data = dat_market,
  spike = prior_spike_slab(pi = 0.1), # CONSERVATIVE
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50),
    omega = list(prior_normal(5, 2, lb=1, ub=12), prior_normal(10, 2, lb=1, ub=12))
  ),
  chains = 2, iter = 2000, warmup = 1000
)
```

#### Option 2: The "Permissive" Model (Capturing the Dip)
Here we use a higher inclusion probability (`pi = 0.8`), allowing the model to capture the finer-grained dip and recovery cycles.

```{r fit-market-perm}
fit_perm <- smoothbp_ss(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker), omega = list(~ ticker, ~ ticker),
  rho = list(~ 1, ~ 1), data = dat_market,
  spike = prior_spike_slab(pi = 0.8), # PERMISSIVE
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50),
    omega = list(prior_normal(5, 2, lb=1, ub=12), prior_normal(10, 2, lb=1, ub=12))
  ),
  chains = 2, iter = 2000, warmup = 1000
)

# Visualize the permissive fit
pred_perm <- fitted(fit_perm)
dat_market$y_fit <- pred_perm$fitted_mean
dat_market$lo    <- pred_perm$fitted_Q2.5
dat_market$hi    <- pred_perm$fitted_Q97.5

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point() +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() +
  labs(title = "Permissive Model (pi = 0.8)", subtitle = "Capturing the structural autumn dip and recovery.")
```

### Analyzing Leadership Across Different Events
By using the permissive model, we can see if the same security "led" both the spring pivot and the autumn dip.

```{r market-prob}
draws <- as_draws_matrix(fit_perm$draws)

# Event 1: AI Pivot
# Prob NVDA led MSFT in the spring
lag1_nvda_msft <- draws[, "omega1_tickerNVDA"] - draws[, "omega1_tickerMSFT"]
p_lead1 <- mean(lag1_nvda_msft < 0)

# Event 2: Year-End Rally
# Prob NVDA led MSFT in the autumn
lag2_nvda_msft <- draws[, "omega2_tickerNVDA"] - draws[, "omega2_tickerMSFT"]
p_lead2 <- mean(lag2_nvda_msft < 0)

message(sprintf("Prob NVDA led AI Pivot: %.2f", p_lead1))
message(sprintf("Prob NVDA led Year-End Rally: %.2f", p_lead2))
```

---

## 3. Shifting Change-Points (Non-stationary Locations)

In many systems, the *location* of a change-point ($\omega$) is not constant but depends on individual characteristics. `smoothbp` allows the change-point itself to have a linear predictor.

### Simulation: Age of onset depends on Covariate X
Imagine a transition that happens earlier for individuals with higher values of $X$.

```{r sim-shifting}
set.seed(456)
n_subj <- 40
n_obs  <- 10
tau    <- seq(0, 10, length.out = n_obs)
x      <- runif(n_subj, 0, 1)

rows <- list()
for (i in 1:n_subj) {
  # omega shifts from 7 down to 3 as x goes from 0 to 1
  om_i <- 7 - 4 * x[i]
  di   <- tau - om_i
  yi   <- 5 + 0.3 * di - 1.2 * di * plogis(di * 5) + rnorm(n_obs, sd = 0.1)
  rows[[i]] <- data.frame(subject = i, tau = tau, y = yi, x = x[i])
}
dat_shift <- do.call(rbind, rows)
```

### Fitting with `omega ~ x`
We specify `omega = list(~ x)` to model this dependency.

```{r fit-shifting}
fit_shift <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),
  omega   = list(~ x),
  deltas  = list(~ 1),
  data    = dat_shift,
  priors  = smoothbp_priors(omega = list(prior_normal(5, 3, lb = 0))),
  chains = 2, iter = 2000, warmup = 1000, seed = 42
)

summary(fit_shift, effects = "fixed") |> filter(grepl("^omega", variable))
```

---

## Conclusion

`smoothbp` provides a unified framework for modeling multiple structural changes that are themselves dependent on other factors. By modeling two breakpoints in financial data, we can reveal the evolving leadership dynamics of complex markets.
