---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 6
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
library(tidyr)
```

---

## 1. Market Turning Points (NASDAQ 2022-2023)

In this example, we model the structural evolution of three tech giants. Rather than simply comparing the "center" of transitions ($\omega$), we derive the **actual turning points**—the local maxima and minima where the price trend reverses.

### The Data
Monthly closing prices from Oct 2022 to Dec 2023.
```{r real-data}
dat_market <- data.frame(
  month  = rep(1:15, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 15),
  price  = c(
    13.50, 16.50, 14.60, 19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    232, 255, 240, 241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    153, 148, 130, 142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)
dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))
```

### Fitting the Model
We use `smoothbp()` with five candidate breakpoints to capture the high-frequency structural shifts of 2023.

```{r fit-market}
fit_market <- smoothbp(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker), 
  omega  = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker),
  rho    = list(~ 1, ~ 1, ~ 1, ~ 1, ~ 1), 
  data   = dat_market,
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50), delta = prior_normal(0, 100),
    omega = list(
      prior_normal(2, 1, lb=1, ub=15), prior_normal(4, 1, lb=1, ub=15),
      prior_normal(7, 1, lb=1, ub=15), prior_normal(10, 1, lb=1, ub=15),
      prior_normal(13, 1, lb=1, ub=15)
    ),
    rho = list(prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1))
  ),
  chains = 4, iter = 4000, warmup = 2000
)
```

### Finding Turning Points via the Analytical Derivative

The `smoothbp` mean function is (from the source of `fitted.smoothbp_fit`):
$$\mu(\tau) = b_0 + b_1(\tau - \omega_1) + \sum_{k=1}^K \delta_k (\tau - \omega_k)\, \sigma(\rho_k(\tau - \omega_k))$$

Applying the product rule to the $k$-th term, letting $d_k = \tau-\omega_k$ and $s_k=\sigma(\rho_k d_k)$:
$$f'(\tau) = b_1 + \sum_{k=1}^K \delta_k \left[s_k + \rho_k d_k s_k(1 - s_k)\right]$$

We solve $f'(\tau) = 0$ numerically: evaluate the fit on a fine grid, compute the slope via finite differences, and find sign-changes.

### Step 1: Turning Points on the Posterior Mean Curve

For **plotting**, we find turning points directly on the posterior mean curve. This avoids the problem of averaging extrema across draws (which produces nonsensical locations when multiple peaks/troughs exist).

```{r turning-points}
grid_months <- seq(1, 15, length.out = 1000)
h <- diff(grid_months)[1]

# Evaluate posterior mean curve on a fine grid, per ticker
pred_fine <- do.call(rbind, lapply(levels(dat_market$ticker), function(tk) {
  nd <- data.frame(month = grid_months, ticker = factor(tk, levels = levels(dat_market$ticker)))
  pf <- fitted(fit_market, newdata = nd)
  data.frame(ticker = tk, month = grid_months, y_fit = pf$fitted_mean)
}))

# Find sign-changes in f'(tau) on the mean curve
find_extrema_on_curve <- function(months, y_fit) {
  dy <- diff(y_fit) / diff(months)
  sc <- which(diff(sign(dy)) != 0)
  if (length(sc) == 0) return(NULL)
  data.frame(
    month     = months[sc + 1],
    y_val     = y_fit[sc + 1],
    direction = ifelse(diff(sign(dy))[sc] < 0, "peak", "trough")
  )
}

extrema_mean <- pred_fine |>
  group_by(ticker) |>
  do(find_extrema_on_curve(.$month, .$y_fit)) |>
  ungroup()

print(extrema_mean)
```

### Visualizing Turning Points

Red markers now sit at the **actual peaks and troughs** of the posterior mean curve — the points where the fitted trend physically reverses direction.

```{r plot-extrema}
pred_orig <- fitted(fit_market)
dat_market$y_fit <- pred_orig$fitted_mean
dat_market$lo    <- pred_orig$fitted_Q2.5
dat_market$hi    <- pred_orig$fitted_Q97.5

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point(alpha = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  geom_point(
    data = extrema_mean,
    aes(x = month, y = y_val, shape = direction),
    color = "red", size = 4, stroke = 1.5, inherit.aes = FALSE
  ) +
  scale_shape_manual(values = c(peak = 4, trough = 3), name = "Turning point") +
  scale_x_continuous(
    breaks = 1:15,
    labels = c("Oct22","Nov","Dec","Jan23","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
  ) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Market Turning Points: where f'(tau) = 0",
    subtitle = "X = peak, + = trough. Located on the posterior mean curve."
  )
```

### Step 2: Posterior Uncertainty Over Turning Points

For **probabilities**, we find extrema on each posterior draw individually, then cluster by proximity to identify corresponding events across tickers.

```{r leadership-robust}
# Per-draw extrema for uncertainty quantification
get_extrema_per_draw <- function(fit, ticker_level, n_draws_use = 500) {
  newdata_fine <- data.frame(
    month  = grid_months,
    ticker = factor(ticker_level, levels = levels(dat_market$ticker))
  )
  fit_raw <- fitted(fit, newdata = newdata_fine, summary = FALSE)
  idx <- sample(nrow(fit_raw), min(n_draws_use, nrow(fit_raw)))
  fit_raw <- fit_raw[idx, , drop = FALSE]

  do.call(rbind, lapply(seq_len(nrow(fit_raw)), function(i) {
    dy <- diff(fit_raw[i, ]) / h
    sc <- which(diff(sign(dy)) != 0)
    if (length(sc) == 0) return(NULL)
    data.frame(
      draw      = i,
      tau_star  = grid_months[sc + 1],
      direction = ifelse(diff(sign(dy))[sc] < 0, "peak", "trough")
    )
  }))
}

extrema_list <- lapply(levels(dat_market$ticker), function(tk) {
  ex <- get_extrema_per_draw(fit_market, tk)
  if (!is.null(ex)) ex$ticker <- tk
  ex
})
extrema_all <- do.call(rbind, extrema_list)

# Cluster peaks by proximity to the mean-curve peaks
# For each draw, find the peak closest to the summer peak (Jul-Aug region)
summer_peak_ref <- extrema_mean |> filter(direction == "peak") |> pull(month) |> max(na.rm = TRUE)

get_nearest_peak <- function(tk, ref) {
  peaks <- extrema_all |> filter(ticker == tk, direction == "peak")
  peaks |>
    mutate(dist = abs(tau_star - ref)) |>
    group_by(draw) |>
    slice_min(dist, n = 1) |>
    ungroup() |>
    pull(tau_star)
}

peak_nvda <- get_nearest_peak("NVDA", summer_peak_ref)
peak_msft <- get_nearest_peak("MSFT", summer_peak_ref)

n_common <- min(length(peak_nvda), length(peak_msft))
p_nvda_first <- mean(peak_nvda[1:n_common] < peak_msft[1:n_common])
message(sprintf("Prob NVDA summer peak occurred before MSFT: %.2f", p_nvda_first))
```

---

## Conclusion

By analyzing the analytical derivative of the `smoothbp` model, we can distinguish between the timing of structural transitions ($\omega$) and the timing of actual market reversals ($f'(\tau)=0$). This provides a mathematically rigorous foundation for lead/lag analysis in non-linear systems.
