---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 4
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
```

---

## 1. Conditional Breakpoints (Emergent Behavior)

A common challenge is when the **existence** of a breakpoint depends on a covariate or experimental group. Instead of fitting separate models, we can use interaction terms in the `deltas` sub-model paired with spike-and-slab priors.

### Simulation: Group-specific structure
```{r fit-conditional}
set.seed(123)
n <- 150
tau <- seq(0, 10, length.out = n)
yA <- 2 + 0.5 * tau + rnorm(n, sd = 0.2)
om <- 5; rho <- 4; delta <- -1.5
di <- tau - om; si <- plogis(di * rho)
yB <- 2 + 0.5 * di + delta * di * si + rnorm(n, sd = 0.2)
dat_cond <- data.frame(y = c(yA, yB), tau = rep(tau, 2), group = rep(c("A", "B"), each = n))

fit_cond <- smoothbp_ss(
  formula = y ~ tau, b0 = ~ group, b1 = ~ group,
  deltas = list(~ group), omega = list(~ 1), rho = list(~ 1),
  data = dat_cond, spike = prior_spike_slab(pi = 0.2, learn_pi = TRUE),
  chains = 2, iter = 2000, warmup = 1000, seed = 42
)

pred_cond <- fitted(fit_cond)
dat_cond$y_fit <- pred_cond$fitted_mean
dat_cond$lo    <- pred_cond$fitted_Q2.5
dat_cond$hi    <- pred_cond$fitted_Q97.5

ggplot(dat_cond, aes(x = tau, y = y, color = group)) +
  geom_point(alpha = 0.2) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = group), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  facet_wrap(~group) + theme_minimal()
```

---

## 2. Market Leader-Follower Dynamics (NASDAQ 2022-2023)

In financial markets, structural shifts often reflect the transition between macro cycles. By including data from late 2022, we can model the transition from the bear market into the AI-driven recovery of 2023.

### The Data: From Bear to Bull
We use monthly data from Oct 2022 to Dec 2023 (15 months).

```{r real-data}
# 15 months: Oct 2022 (1) to Dec 2023 (15)
dat_market <- data.frame(
  month  = rep(1:15, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 15),
  price  = c(
    # NVDA
    13.50, 16.50, 14.60, 19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    # MSFT
    232, 255, 240, 241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    # AAPL
    153, 148, 130, 142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)

dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))
```

### High-Capacity Structural Modeling
To capture the full sequence of events (Post-Bear Pivot, March Dip, AI Pivot, Autumn Dip, and Year-End Recovery), we use **five candidate breakpoints** and `smoothbp()` for efficient discovery.

```{r fit-market-perm}
fit_perm <- smoothbp(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker), 
  omega  = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker),
  rho    = list(~ 1, ~ 1, ~ 1, ~ 1, ~ 1), 
  data   = dat_market,
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50), delta = prior_normal(0, 100),
    omega = list(
      prior_normal(2, 1, lb=1, ub=15), prior_normal(5, 1, lb=1, ub=15),
      prior_normal(8, 1, lb=1, ub=15), prior_normal(12, 1, lb=1, ub=15),
      prior_normal(14, 1, lb=1, ub=15)
    ),
    rho = list(prior_normal(15, 5, lb=1), prior_normal(15, 5, lb=1), prior_normal(15, 5, lb=1), prior_normal(15, 5, lb=1), prior_normal(15, 5, lb=1))
  ),
  chains = 4, iter = 5000, warmup = 3000
)

# Visualize with monthly breaks
pred_perm <- fitted(fit_perm)
dat_market$y_fit <- pred_perm$fitted_mean
dat_market$lo    <- pred_perm$fitted_Q2.5
dat_market$hi    <- pred_perm$fitted_Q97.5

ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point() +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  scale_x_continuous(breaks = 1:15, labels = c("Oct22", "Nov", "Dec", "Jan23", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Full Structural Narrative (15 Months, 5 BP)", subtitle = "Capturing the transition from Bear to AI Bull.")
```

### Leadership Across Macro Events
We can now precisely quantify who led each of the five major market phases.

```{r market-prob}
draws <- as_draws_matrix(fit_perm$draws)
# Example: AI Pivot (BP 3)
p_lead_ai <- mean((draws[, "omega3_tickerNVDA"] - draws[, "omega3_tickerMSFT"]) < 0)
message(sprintf("Prob NVDA led the AI Pivot (Event 3): %.2f", p_lead_ai))
```

---

## 3. Shifting Change-Points (Non-stationary Locations)

`smoothbp` allows the change-point location ($\omega$) to depend on covariates.

```{r fit-shifting}
set.seed(456)
n_subj <- 40; n_obs <- 10; tau <- seq(0, 10, length.out = n_obs); x <- runif(n_subj, 0, 1)
rows <- list()
for (i in 1:n_subj) {
  om_i <- 7 - 4 * x[i]; di <- tau - om_i
  yi   <- 5 + 0.3 * di - 1.2 * di * plogis(di * 5) + rnorm(n_obs, sd = 0.1)
  rows[[i]] <- data.frame(subject = i, tau = tau, y = yi, x = x[i])
}
dat_shift <- do.call(rbind, rows)

fit_shift <- smoothbp(
  formula = y ~ tau, b0 = ~ 1 + (1 | subject), omega = list(~ x), deltas = list(~ 1),
  data = dat_shift, priors = smoothbp_priors(omega = list(prior_normal(5, 3, lb = 0))),
  chains = 2, iter = 2000, warmup = 1000, seed = 42
)
summary(fit_shift, effects = "fixed") |> filter(grepl("^omega", variable))
```

---

## Conclusion

By treating $\delta$, $\omega$, and $\rho$ as parameters that can vary with predictors, `smoothbp` provides a unified framework for modeling structural changes in complex systems.
