---
title: "Advanced Modeling with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Modeling with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 6
)
```

While `smoothbp` is efficient for standard piecewise regression, its real power lies in its ability to handle structural changes that are themselves functions of other variables. This vignette demonstrates "edge cases" where `smoothbp` offers flexibility that is difficult to achieve with other packages.

```{r libs, message=FALSE, warning=FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..")
} else {
  library(smoothbp)
}
library(ggplot2)
library(dplyr)
library(posterior)
library(tidyr)
```

---

## 1. Market Turning Points (NASDAQ 2022-2023)

In this example, we model the structural evolution of three tech giants. Rather than simply comparing the "center" of transitions ($\omega$), we derive the **actual turning points**—the local maxima and minima where the price trend reverses.

### The Data
Monthly closing prices from Oct 2022 to Dec 2023.
```{r real-data}
dat_market <- data.frame(
  month  = rep(1:15, 3),
  ticker = rep(c("NVDA", "MSFT", "AAPL"), each = 15),
  price  = c(
    13.50, 16.50, 14.60, 19.52, 23.19, 27.75, 27.72, 37.80, 42.27, 46.69, 49.32, 43.47, 40.75, 46.74, 49.49,
    232, 255, 240, 241.47, 243.65, 281.63, 300.15, 321.49, 333.39, 328.87, 321.56, 309.77, 331.71, 372.49, 369.67,
    153, 148, 130, 142.01, 145.30, 162.55, 167.26, 174.96, 191.46, 193.91, 185.69, 169.23, 168.79, 188.00, 190.55
  )
)
dat_market$ticker <- factor(dat_market$ticker, levels = c("AAPL", "MSFT", "NVDA"))
```

### Fitting the Model
We use `smoothbp()` with five candidate breakpoints to capture the high-frequency structural shifts of 2023.

```{r fit-market}
fit_market <- smoothbp(
  formula = price ~ month, b0 = ~ ticker,
  deltas = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker), 
  omega  = list(~ ticker, ~ ticker, ~ ticker, ~ ticker, ~ ticker),
  rho    = list(~ 1, ~ 1, ~ 1, ~ 1, ~ 1), 
  data   = dat_market,
  priors = smoothbp_priors(
    b0 = prior_normal(200, 500), b1 = prior_normal(0, 50), delta = prior_normal(0, 100),
    omega = list(
      prior_normal(2, 1, lb=1, ub=15), prior_normal(4, 1, lb=1, ub=15),
      prior_normal(7, 1, lb=1, ub=15), prior_normal(10, 1, lb=1, ub=15),
      prior_normal(13, 1, lb=1, ub=15)
    ),
    rho = list(prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1), prior_normal(20, 5, lb=1))
  ),
  chains = 4, iter = 4000, warmup = 2000
)
```

### Finding Turning Points via the Analytical Derivative

The `smoothbp` mean function is (from the source of `fitted.smoothbp_fit`):
$$\mu(\tau) = b_0 + b_1(\tau - \omega_1) + \sum_{k=1}^K \delta_k (\tau - \omega_k)\, \sigma(\rho_k(\tau - \omega_k))$$

Applying the product rule to the $k$-th term, letting $d_k = \tau-\omega_k$ and $s_k=\sigma(\rho_k d_k)$:
$$f'(\tau) = b_1 + \sum_{k=1}^K \delta_k \left[s_k + \rho_k d_k s_k(1 - s_k)\right]$$

The equation $f'(\tau) = 0$ is transcendental (no closed form), but we can evaluate the derivative **analytically** and use `uniroot()` near each $\omega_k$ to find the associated turning point. This directly ties each turning point to its causal breakpoint.

```{r turning-points}
n_bp <- 5
draws <- as_draws_matrix(fit_market$draws)
tickers <- levels(dat_market$ticker) # AAPL (baseline), MSFT, NVDA

# Extract ticker-specific parameters for a single draw
get_params <- function(d, tk) {
  # Helper to safely extract a column value
  val <- function(pat) {
    v <- d[grep(paste0("^", pat, "$"), names(d))]
    if (length(v) == 0) return(0)
    unname(v[1])
  }
  
  b1 <- val("b1_\\(Intercept\\)")
  if (tk != tickers[1]) b1 <- b1 + val(paste0("b1_ticker", tk))
  
  deltas <- omegas <- rhos <- numeric(n_bp)
  for (k in 1:n_bp) {
    deltas[k] <- val(paste0("delta", k, "_\\(Intercept\\)"))
    omegas[k] <- val(paste0("omega", k, "_\\(Intercept\\)"))
    rhos[k]   <- val(paste0("rho", k, "_\\(Intercept\\)"))
    
    if (tk != tickers[1]) {
      deltas[k] <- deltas[k] + val(paste0("delta", k, "_ticker", tk))
      omegas[k] <- omegas[k] + val(paste0("omega", k, "_ticker", tk))
      rhos[k]   <- rhos[k]   + val(paste0("rho", k, "_ticker", tk))
    }
  }
  list(b1 = b1, deltas = deltas, omegas = omegas, rhos = rhos)
}

# Analytical derivative
fprime <- function(tau, p) {
  if (is.na(tau)) return(NA)
  slope <- p$b1
  for (k in 1:n_bp) {
    dk <- tau - p$omegas[k]
    sk <- plogis(dk * p$rhos[k])
    slope <- slope + p$deltas[k] * (sk + p$rhos[k] * dk * sk * (1 - sk))
  }
  slope
}

# For each bp k, search for f'(tau)=0 near omega_k
find_turning_near_omega <- function(p, window = 2) {
  out <- list()
  for (k in 1:n_bp) {
    if (is.na(p$omegas[k])) next
    lo <- max(1, p$omegas[k] - window)
    hi <- min(15, p$omegas[k] + window)
    f_lo <- fprime(lo, p)
    f_hi <- fprime(hi, p)
    
    if (is.na(f_lo) || is.na(f_hi)) next
    if (f_lo * f_hi < 0) {
      root <- try(uniroot(fprime, c(lo, hi), p = p)$root, silent = TRUE)
      if (!inherits(root, "try-error")) {
        out[[length(out) + 1]] <- data.frame(
          bp = k, tau_star = root,
          direction = ifelse(f_lo > 0, "peak", "trough")
        )
      }
    }
  }
  if (length(out) == 0) return(NULL)
  do.call(rbind, out)
}
```

### Turning Points on the Posterior Mean

```{r mean-turning-points}
# Use posterior means of all parameters
d_mean <- colMeans(draws)

tp_mean <- do.call(rbind, lapply(tickers, function(tk) {
  p <- get_params(d_mean, tk)
  tp <- find_turning_near_omega(p)
  if (!is.null(tp)) tp$ticker <- tk
  tp
}))

# Get y-values from the fitted curve
pred_orig <- fitted(fit_market)
dat_market$y_fit <- pred_orig$fitted_mean
dat_market$lo    <- pred_orig$fitted_Q2.5
dat_market$hi    <- pred_orig$fitted_Q97.5

tp_mean$y_val <- sapply(1:nrow(tp_mean), function(i) {
  idx <- dat_market$ticker == tp_mean$ticker[i]
  approx(dat_market$month[idx], dat_market$y_fit[idx], xout = tp_mean$tau_star[i])$y
})

print(tp_mean)
```

### Visualizing Turning Points

```{r plot-extrema}
ggplot(dat_market, aes(x = month, y = price, color = ticker)) +
  geom_point(alpha = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = ticker), alpha = 0.1, color = NA) +
  geom_line(aes(y = y_fit), size = 1) +
  geom_point(
    data = tp_mean, aes(x = tau_star, y = y_val, shape = direction),
    color = "red", size = 4, stroke = 1.5, inherit.aes = FALSE
  ) +
  scale_shape_manual(values = c(peak = 4, trough = 3), name = "Turning point") +
  scale_x_continuous(breaks = 1:15, labels = c(
    "Oct22","Nov","Dec","Jan23","Feb","Mar","Apr","May",
    "Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  facet_wrap(~ticker, scales = "free_y") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Market Turning Points (semi-analytical)",
       subtitle = "Solving f'(tau) = 0 near each omega_k. X = peak, + = trough.")
```

### Posterior Uncertainty and Leadership

For each draw, we solve $f'(\tau)=0$ near each $\omega_k$, giving a full posterior over turning point timing.

```{r leadership-robust}
# Sample draws for speed
set.seed(42)
idx <- sample(nrow(draws), min(500, nrow(draws)))

tp_draws_list <- lapply(idx, function(i) {
  d <- draws[i, ]
  do.call(rbind, lapply(tickers, function(tk) {
    p <- get_params(d, tk)
    tp <- find_turning_near_omega(p)
    if (!is.null(tp)) { tp$ticker <- tk; tp$draw <- i }
    tp
  }))
})
tp_draws <- do.call(rbind, tp_draws_list)

# Compare the first peak (summer) across tickers
# Use bp identity to ensure we compare the same structural event
if (!is.null(tp_mean) && !is.null(tp_draws)) {
  summer_bp_df <- tp_mean |> filter(direction == "peak")
  if (nrow(summer_bp_df) > 0) {
    summer_bp <- summer_bp_df |> 
      group_by(ticker) |> slice_max(tau_star) |> pull(bp) |> unique()
    
    if (length(summer_bp) > 0) {
      ref_bp <- summer_bp[1]
      peak_nvda <- tp_draws |> filter(ticker == "NVDA", bp == ref_bp) |> pull(tau_star)
      peak_msft <- tp_draws |> filter(ticker == "MSFT", bp == ref_bp) |> pull(tau_star)
      n_common <- min(length(peak_nvda), length(peak_msft))
      if (n_common > 10) { # Ensure we have enough draws for a stable estimate
        p_lead <- mean(peak_nvda[1:n_common] < peak_msft[1:n_common])
        message(sprintf("Prob NVDA summer peak before MSFT (BP %d): %.2f", ref_bp, p_lead))
      } else {
        message("Not enough posterior draws found a turning point for this event.")
      }
    }
  } else {
    message("No peaks found on the posterior mean curve.")
  }
} else {
  message("No turning points discovered in the posterior sample.")
}
```

---

## Conclusion

By analyzing the analytical derivative of the `smoothbp` model, we can distinguish between the timing of structural transitions ($\omega$) and the timing of actual market reversals ($f'(\tau)=0$). This provides a mathematically rigorous foundation for lead/lag analysis in non-linear systems.
