---
title: "Validating smoothbp against brms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating smoothbp against brms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 96
)
```

## Why this vignette exists

`smoothbp` is a bespoke Metropolis-within-Gibbs sampler for one specific
model: a hierarchical piecewise regression with a logistic-smoothed
change-point. The recovery test in `tests/testthat/test-smoothbp.R` shows
the sampler returns posterior intervals that contain the data-generating
values, which is necessary, but it is a self-consistency check: simulator
and likelihood share the same code path. To trust the package one also
wants to see that it agrees with an independent implementation. `brms`
(via Stan) is the natural reference, fitting the same generative model
via NUTS on a custom non-linear formula.

### One important practical note before we start

When the change-point parameter `omega` drifts outside the range of the
time variable, the smoothing term collapses and `b2` becomes effectively
unidentified. NUTS chains that initialise in that region can get stuck
there for the entire run, producing an artificially multimodal posterior
that has nothing to do with the data. The clean fix is to bound `omega`
above at `max(tau)` in both samplers, and to initialise brms chains near
the prior mean (which is what smoothbp does internally). Both fixes
applied below.

```{r setup, message = FALSE, warning = FALSE}
library(smoothbp)
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)

have_brms <- requireNamespace("brms", quietly = TRUE)
if (!have_brms) {
  message("brms not installed -- install.packages('brms') to render the comparison.")
}
```

## Simulate one dataset

```{r simulate}
set.seed(31)
dat <- simulate_smoothbp(
  n_subj = 25, n_obs = 10,
  b0 = 5.0, b1 = -0.4, b2 = 1.4,
  omega = 3.2, rho = 4.0,
  sigma = 0.4, sigma_u = 0.7,
  seed = 31L
)
true_params(dat)
```

## Fit with smoothbp

```{r fit-smoothbp}
t0 <- Sys.time()
fit_sbp <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),
  b1      = ~ 1,
  b2      = ~ 1,
  omega   = ~ 1,
  rho     = ~ 1,
  data    = dat,
  priors  = smoothbp_priors(
    b0      = prior_normal(0, 10),
    b1      = prior_normal(0, 2),
    b2      = prior_normal(0, 2),
    omega   = prior_normal(3, 2, lb = 0, ub = max(dat$tau)),
    rho     = prior_normal(3, 2, lb = 0)
  ),
  chains  = 4L, iter = 2000L, warmup = 1000L,
  seed    = 31L, .verbose = FALSE
)
time_sbp <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
```

## Fit with brms (Stan / NUTS)

```{r fit-brms, eval = have_brms, message = FALSE, warning = FALSE, results = "hide"}
suppressPackageStartupMessages(library(brms))

bf_smoothed <- brms::bf(
  y ~ b0 + b1 * (tau - omega) +
        b2 * (tau - omega) / (1 + exp(-(tau - omega) * rho)),
  b0 ~ 1 + (1 | subject),
  b1 ~ 1, b2 ~ 1, omega ~ 1, rho ~ 1,
  nl = TRUE
)

# brms::prior() captures arguments by non-standard evaluation, so
# `ub = max(dat$tau)` would be embedded literally in the generated Stan
# code. Pre-compute the value and pass it via prior_string() which uses
# standard evaluation and accepts numeric bounds.
ub_omega <- max(dat$tau)

priors_brms <- c(
  brms::prior(normal(0, 10), nlpar = "b0"),
  brms::prior(normal(0, 2),  nlpar = "b1"),
  brms::prior(normal(0, 2),  nlpar = "b2"),
  brms::prior_string("normal(3, 2)", nlpar = "omega",
                     lb = 0, ub = ub_omega),
  brms::prior(normal(3, 2),  nlpar = "rho",   lb = 0)
)

# Initialise every chain near the prior mean.  Without this, Stan's default
# uniform-on-(-2, 2) initialisation lands some chains in the unidentifiable
# omega > max(tau) trap and they never escape.
init_fun <- function() list(
  b_b0    = array(rnorm(1, 5, 1)),
  b_b1    = array(rnorm(1, 0, 0.3)),
  b_b2    = array(rnorm(1, 0, 0.3)),
  b_omega = array(rnorm(1, 3, 0.3)),
  b_rho   = array(rnorm(1, 3, 0.5))
)

t0 <- Sys.time()
fit_brms <- brms::brm(
  bf_smoothed,
  data    = dat,
  prior   = priors_brms,
  chains  = 4, iter = 2000, warmup = 1000,
  seed    = 31, refresh = 0,
  init    = init_fun,
  control = list(adapt_delta = 0.95)
)
time_brms <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
```

## Did brms converge?

```{r brms-convergence, eval = have_brms}
sbp_names <- c(
  "b0_(Intercept)", "b1_(Intercept)", "b2_(Intercept)",
  "omega_(Intercept)", "rho_(Intercept)",
  "sigma", "sigma_u"
)
brms_names <- c(
  "b_b0_Intercept", "b_b1_Intercept", "b_b2_Intercept",
  "b_omega_Intercept", "b_rho_Intercept",
  "sigma", "sd_subject__b0_Intercept"
)

posterior::summarise_draws(
  posterior::as_draws_df(fit_brms), rhat, ess_bulk
) %>%
  dplyr::filter(variable %in% brms_names) %>%
  dplyr::transmute(parameter = sbp_names[match(variable, brms_names)],
                   rhat, ess_bulk) %>%
  knitr::kable(digits = 3,
               caption = "brms convergence on population-level parameters.")
```

If any `rhat` exceeds 1.05, brms has not mixed and the comparison below is
contaminated; do not interpret disagreements as smoothbp errors. With the
bounded `omega` prior and the seeded `init_fun`, all four chains should land
in the same basin.

## Compare posterior summaries

```{r compare-summaries, eval = have_brms}
sbp_draws  <- posterior::as_draws_df(fit_sbp$draws)[, sbp_names]
brms_draws <- as.data.frame(posterior::as_draws_df(fit_brms))[, brms_names]
colnames(brms_draws) <- sbp_names

draws_long <- dplyr::bind_rows(
  tidyr::pivot_longer(sbp_draws,  cols = dplyr::everything(),
                      names_to = "parameter", values_to = "value") %>%
    dplyr::mutate(package = "smoothbp"),
  tidyr::pivot_longer(brms_draws, cols = dplyr::everything(),
                      names_to = "parameter", values_to = "value") %>%
    dplyr::mutate(package = "brms")
)

draws_long %>%
  dplyr::group_by(parameter, package) %>%
  dplyr::summarise(
    mean = mean(value),
    sd   = sd(value),
    q025 = quantile(value, 0.025),
    q975 = quantile(value, 0.975),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from  = package,
    values_from = c(mean, sd, q025, q975),
    names_glue  = "{package}_{.value}"
  ) %>%
  dplyr::mutate(
    delta_mean_pct = 100 * abs(smoothbp_mean - brms_mean) /
                       (abs(brms_mean) + 1e-8)
  ) %>%
  knitr::kable(digits = 3,
               caption = "Posterior summaries for both packages.")
```

## Overlaid marginal posteriors

```{r overlay, eval = have_brms, fig.height = 5}
ggplot(draws_long, aes(x = value, fill = package, colour = package)) +
  geom_density(alpha = 0.35) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values  = c(smoothbp = "#1f77b4", brms = "#d62728")) +
  scale_colour_manual(values = c(smoothbp = "#1f77b4", brms = "#d62728")) +
  labs(
    title    = "Marginal posteriors: smoothbp vs brms",
    subtitle = sprintf(
      "Same simulated data, same priors. smoothbp: %.1fs   brms: %.1fs",
      time_sbp, time_brms
    ),
    x = NULL, y = "density"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Effective sample size per second

```{r ess, eval = have_brms}
ess_sbp  <- posterior::summarise_draws(fit_sbp$draws, ess_bulk) %>%
  dplyr::filter(variable %in% sbp_names) %>%
  dplyr::transmute(parameter = variable,
                   smoothbp_ess_per_sec = ess_bulk / time_sbp)

ess_brms <- posterior::summarise_draws(
  posterior::as_draws_df(fit_brms), ess_bulk
) %>%
  dplyr::filter(variable %in% brms_names) %>%
  dplyr::mutate(parameter = sbp_names[match(variable, brms_names)]) %>%
  dplyr::transmute(parameter, brms_ess_per_sec = ess_bulk / time_brms)

dplyr::full_join(ess_sbp, ess_brms, by = "parameter") %>%
  knitr::kable(digits = 1,
               caption = "Bulk ESS / second by package and parameter.")
```

## What to look for

If both samplers have converged, the marginal posteriors overlay almost
exactly, the means agree to within Monte Carlo error (`delta_mean_pct`
should be a fraction of a per cent), and the credible intervals are within
sampling variability of each other. ESS per second is substantially higher
for `smoothbp` because each iteration is cheaper, even though NUTS gets
more out of each iteration.

If you remove the `ub = max(dat$tau)` bound or the `init_fun`, you will
likely see the brms posterior on `omega` and `b2` go bimodal: one cluster
near the truth, and a second cluster in the unidentifiable region with
`omega > max(tau)` and `b2` pinned at a ridiculous value matching the
right-edge slope. That is a sampler-mixing artefact, not a real feature of
the posterior, and worth demonstrating in your own analyses before trusting
results from change-point models that allow `omega` to escape the data
range.

A note on smoothbp's robustness: smoothbp initialises every chain near the
prior mean, which keeps it out of the degenerate region by construction.
That is excellent in practice when the user has a sensible prior on
`omega`, and is the typical case. It also means that on a problem where
the posterior is genuinely multimodal, the chains may all converge to
whichever mode is closest to the prior mean and fail to explore the
others. If you suspect multimodality, fit with deliberately overdispersed
initial values by varying the prior across runs.
