---
title: "Validating smoothbp against brms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating smoothbp against brms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 96
)
```

## Why this vignette exists

`smoothbp` is a bespoke Metropolis-within-Gibbs sampler, written in Rust, for
one specific model: a hierarchical piecewise regression with a logistic-smoothed
change-point. The recovery test in `tests/testthat/test-smoothbp.R` shows the
sampler returns posterior intervals that contain the data-generating values,
which is necessary, but it is a self-consistency check, simulator and likelihood
share the same code path. To trust the package for inference one also wants to
see that it agrees with an independent implementation. `brms` (and Stan
underneath it) is the natural reference because it can fit the exact same
generative model via NUTS on a custom non-linear formula.

This vignette fits the same intercept-only data with both packages and overlays
the marginal posteriors. Identical models, identical priors, comparable wall
clock, no Bayes factor sleight of hand.

```{r setup, message = FALSE, warning = FALSE}
library(smoothbp)
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)

have_brms <- requireNamespace("brms", quietly = TRUE)
if (!have_brms) {
  message("brms not installed -- install.packages('brms') to render the comparison.")
}
```

## Simulate one dataset

We use the same generative model the package fits, with no covariates in any
component so the brms parameterisation stays compact.

```{r simulate}
set.seed(31)
dat <- simulate_smoothbp(
  n_subj = 25, n_obs = 10,
  b0 = 5.0, b1 = -0.4, b2 = 1.4,
  omega = 3.2, rho = 4.0,
  sigma = 0.4, sigma_u = 0.7,
  seed = 31L
)
true_params(dat)
```

## Fit with smoothbp

```{r fit-smoothbp}
t0 <- Sys.time()
fit_sbp <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),
  b1      = ~ 1,
  b2      = ~ 1,
  omega   = ~ 1,
  rho     = ~ 1,
  data    = dat,
  priors  = smoothbp_priors(
    b0      = prior_normal(0, 10),
    b1      = prior_normal(0, 2),
    b2      = prior_normal(0, 2),
    omega   = prior_normal(3, 2, lb = 0),
    rho     = prior_normal(3, 2, lb = 0)
  ),
  chains  = 4L, iter = 2000L, warmup = 1000L,
  seed    = 31L, .verbose = FALSE
)
time_sbp <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
```

## Fit with brms (Stan / NUTS)

The model is written as a non-linear `brms` formula matching the `smoothbp`
specification. Priors are deliberately the same as above so the two posteriors
should differ only by Monte Carlo error.

```{r fit-brms, eval = have_brms, message = FALSE, warning = FALSE, results = "hide"}
suppressPackageStartupMessages(library(brms))

bf_smoothed <- brms::bf(
  y ~ b0 + b1 * (tau - omega) +
        b2 * (tau - omega) / (1 + exp(-(tau - omega) * rho)),
  b0 ~ 1 + (1 | subject),
  b1 ~ 1,
  b2 ~ 1,
  omega ~ 1,
  rho ~ 1,
  nl = TRUE
)

priors_brms <- c(
  brms::prior(normal(0, 10), nlpar = "b0"),
  brms::prior(normal(0, 2),  nlpar = "b1"),
  brms::prior(normal(0, 2),  nlpar = "b2"),
  brms::prior(normal(3, 2),  nlpar = "omega", lb = 0),
  brms::prior(normal(3, 2),  nlpar = "rho",   lb = 0)
)

t0 <- Sys.time()
fit_brms <- brms::brm(
  bf_smoothed,
  data    = dat,
  prior   = priors_brms,
  chains  = 4, iter = 2000, warmup = 1000,
  seed    = 31, refresh = 0,
  control = list(adapt_delta = 0.95)
)
time_brms <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
```

## Compare posterior summaries

```{r compare-summaries, eval = have_brms}
# Pull marginal posteriors for the five population-level scalars and the
# two variance components.
sbp_names <- c(
  "b0_(Intercept)", "b1_(Intercept)", "b2_(Intercept)",
  "omega_(Intercept)", "rho_(Intercept)",
  "sigma", "sigma_u"
)
brms_names <- c(
  "b_b0_Intercept", "b_b1_Intercept", "b_b2_Intercept",
  "b_omega_Intercept", "b_rho_Intercept",
  "sigma", "sd_subject__b0_Intercept"
)

sbp_draws  <- posterior::as_draws_df(fit_sbp$draws)[, sbp_names]
brms_draws <- as.data.frame(posterior::as_draws_df(fit_brms))[, brms_names]
colnames(brms_draws) <- sbp_names

draws_long <- dplyr::bind_rows(
  tidyr::pivot_longer(sbp_draws,  cols = dplyr::everything(),
                      names_to = "parameter", values_to = "value") %>%
    dplyr::mutate(package = "smoothbp"),
  tidyr::pivot_longer(brms_draws, cols = dplyr::everything(),
                      names_to = "parameter", values_to = "value") %>%
    dplyr::mutate(package = "brms")
)

summary_tbl <- draws_long %>%
  dplyr::group_by(parameter, package) %>%
  dplyr::summarise(
    mean = mean(value),
    sd   = sd(value),
    q025 = quantile(value, 0.025),
    q975 = quantile(value, 0.975),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from  = package,
    values_from = c(mean, sd, q025, q975),
    names_glue  = "{package}_{.value}"
  ) %>%
  dplyr::mutate(
    delta_mean_pct = 100 * abs(smoothbp_mean - brms_mean) /
                       (abs(brms_mean) + 1e-8)
  )

knitr::kable(summary_tbl, digits = 3,
             caption = "Posterior summaries for both packages on the same data.")
```

## Overlaid marginal posteriors

```{r overlay, eval = have_brms, fig.height = 5}
ggplot(draws_long, aes(x = value, fill = package, colour = package)) +
  geom_density(alpha = 0.35) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values  = c(smoothbp = "#1f77b4", brms = "#d62728")) +
  scale_colour_manual(values = c(smoothbp = "#1f77b4", brms = "#d62728")) +
  labs(
    title    = "Marginal posteriors: smoothbp vs brms",
    subtitle = sprintf(
      "Same simulated data, same priors. smoothbp: %.1fs   brms: %.1fs",
      time_sbp, time_brms
    ),
    x = NULL, y = "density"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Effective sample size per second

ESS / second is the right quantity for comparing samplers, since smoothbp
trades NUTS's per-iteration sophistication for many more iterations per
second.

```{r ess, eval = have_brms}
ess_sbp  <- posterior::summarise_draws(fit_sbp$draws, ess_bulk) %>%
  dplyr::filter(variable %in% sbp_names) %>%
  dplyr::transmute(parameter = variable,
                   smoothbp_ess_per_sec = ess_bulk / time_sbp)

# Pull bulk ESS from the brms fit, matching the column subset.
ess_brms_full <- posterior::summarise_draws(
  posterior::as_draws_df(fit_brms), ess_bulk
)
ess_brms <- ess_brms_full %>%
  dplyr::filter(variable %in% brms_names) %>%
  dplyr::mutate(parameter = sbp_names[match(variable, brms_names)]) %>%
  dplyr::transmute(parameter, brms_ess_per_sec = ess_bulk / time_brms)

dplyr::full_join(ess_sbp, ess_brms, by = "parameter") %>%
  knitr::kable(digits = 1,
               caption = "Bulk ESS / second by package and parameter.")
```

## What to look for

If `smoothbp` is doing its job, the marginal posteriors overlay almost
exactly, the posterior means agree to within Monte Carlo error (`delta_mean_pct`
should be a fraction of a per cent on most parameters), and the credible
intervals are within sampling variability of each other. ESS per second
should be substantially higher for `smoothbp` because each iteration is
cheaper, even though NUTS gets more out of each iteration.

Disagreement on a parameter, especially `omega`, would be a signal to dig:
either the smoothbp sampler is misspecified, the priors are not actually
identical (truncation in particular is easy to mis-encode), or one of the
chains has not mixed and the comparison is contaminated by Monte Carlo noise.
