---
title: "Addressing Structural Differences with fibr Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Addressing Structural Differences with fibr Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The Problem: Group-Level Structural Differences

In multi-subject change-point models, we often fit a common set of candidate breakpoints across all participants. However, different subjects may have different numbers of actual change-points. For example:
* **Group A**: Participants have two structural breaks (2 knots).
* **Group B**: Participants have only one structural break (1 knot).

If we fit a 2-breakpoint model to all participants:
1. For **Group A**, the second changepoint location ($\omega_2$) is well-defined by the data because there is a non-zero slope change ($\delta_2 \neq 0$).
2. For **Group B**, the second slope change is zero ($\delta_2 = 0$). This makes the location of the second changepoint ($\omega_2$) completely unidentified by the data.

When a parameter is unidentified, its subject-level random deviation ($u_{\omega_2}$) collapses to zero as the group-level standard deviation ($\sigma_{re, \omega_2}$) goes to zero, creating a **hierarchical funnel** (Neal's Funnel). Under a standard centred parameterisation, Hamiltonian Monte Carlo (HMC) struggles in this funnel, resulting in divergent transitions, poor mixing, and flat gradients that can prevent the sampler from finding the true parameters.

This vignette demonstrates how `fibr` diagnoses these structural differences, and how to use `reparameterise = "omega"` to apply full non-centring and restore sampler efficiency.

---

## Step 1: Simulate the Data

We simulate $N = 20$ participants:
* **Subjects 1 to 10**: Two obvious breakpoints ($\delta_1 = 2.0$ at $\omega_1 = 2.0$, and $\delta_2 = -2.0$ at $\omega_2 = 4.0$).
* **Subjects 11 to 20**: Only one obvious breakpoint ($\delta_1 = 2.0$ at $\omega_1 = 2.0$, and $\delta_2 = 0.0$ at $\omega_2 = 4.0$).

```{r simulate-data, message=FALSE, warning=FALSE}
library(smoothbp)
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)
library(brms)

# Simulate subjects 1-10 (2 knots)
dat1 <- simulate_smoothbp(
  n_subj    = 10L,
  n_obs     = 15L,
  b0        = 5.0,
  b1        = -1.0,
  delta     = c(2.0, -2.0),
  omega     = c(2.0, 4.0),
  rho       = c(4.0, 4.0),
  sigma     = 0.15,
  sigma_u   = 0.0,
  tau_range = c(0.5, 5.5),
  seed      = 123L
)
dat1$subject <- as.integer(as.character(dat1$subject))

# Simulate subjects 11-20 (1 knot)
dat2 <- simulate_smoothbp(
  n_subj    = 10L,
  n_obs     = 15L,
  b0        = 5.0,
  b1        = -1.0,
  delta     = c(2.0, 0.0),
  omega     = c(2.0, 4.0),
  rho       = c(4.0, 4.0),
  sigma     = 0.15,
  sigma_u   = 0.0,
  tau_range = c(0.5, 5.5),
  seed      = 456L
)
dat2$subject <- as.integer(as.character(dat2$subject)) + 10L

# Combine
dat <- rbind(dat1, dat2)
dat$subject <- factor(dat$subject)
```

Let's visualize the differences between a 2-knot subject (Subject 1) and a 1-knot subject (Subject 11):

```{r plot-raw-data, fig.width=7, fig.height=4}
example_dat <- dat %>% 
  filter(subject %in% c(1, 11)) %>%
  mutate(Knots = factor(ifelse(subject == 1, "2 Knots (Subject 1)", "1 Knot (Subject 11)")))

ggplot(example_dat, aes(x = tau, y = y, color = Knots)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_line(aes(y = mu), linewidth = 1) +
  labs(
    title = "Structural Differences between Groups",
    subtitle = "Solid lines show the true underlying trajectories",
    x = "Time (tau)", y = "Response (y)"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

---

## Step 2: Fit the Naive Centred Model

We fit a 2-breakpoint `smoothbp_ss` model with a standard centred parameterisation (`reparameterise = "none"`), timing the execution:

```{r fit-centred, message=FALSE}
t0 <- Sys.time()
fit_centred <- smoothbp_ss(
  formula = y ~ tau,
  b0      = ~ 1,
  b1      = ~ 1,
  deltas  = list(~ 1 + (1|subject), ~ 0 + (1|subject)),
  omega   = list(~ 1 + (1|subject), ~ 1 + (1|subject)),
  rho     = list(~ 1, ~ 1),
  data    = dat,
  priors  = smoothbp_priors(
    omega = prior_normal(3.0, 1.5, lb = 0.0)
  ),
  spike   = prior_spike_slab(pi = 0.5),
  reparameterise = "none",
  chains  = 2L,
  iter    = 2000L,
  warmup  = 1000L,
  seed    = 777L,
  .verbose = FALSE
)
t_centred <- difftime(Sys.time(), t0, units = "secs")
```

---

## Step 3: Visualising the Funnel Geometry

When using a centred parameterisation, subject-level random effects are directly modeled as $u_{s} \sim N(0, \sigma^2_{re})$.
* For **Subject 1 (2 knots)**: The location $\omega_2$ is well-identified, so $u_1$ remains clustered around its true value even when the group standard deviation $\sigma_{re}$ is small.
* For **Subject 11 (1 knot)**: The location is unidentified, so $u_{11}$ collapses tightly to 0 as $\sigma_{re}$ decreases. As $\sigma_{re}$ increases, $u_{11}$ can take wider values. This creates a classic **funnel shape** (Neal's Funnel).

We can visualize this funnel geometry by plotting the subject-specific random effects against the log of the group standard deviation:

```{r plot-funnel, fig.width=7, fig.height=4}
draws_centred <- as_draws_df(fit_centred$draws)

df_funnel <- data.frame(
  log_sigma_re = log(draws_centred$sigma_re_omega2),
  u_subj_1     = draws_centred$omega2_re_subject_1,  # subject 1
  u_subj_11    = draws_centred$omega2_re_subject_11  # subject 11
)

df_funnel_long <- df_funnel %>%
  pivot_longer(cols = c(u_subj_1, u_subj_11), names_to = "Subject", values_to = "u") %>%
  mutate(Subject = factor(ifelse(Subject == "u_subj_1", "Subject 1 (2 Knots - Identified)", "Subject 11 (1 Knot - Unidentified)")))

ggplot(df_funnel_long, aes(x = u, y = log_sigma_re, color = Subject)) +
  geom_point(alpha = 0.4, size = 1.2) +
  facet_wrap(~Subject, scales = "free_x") +
  labs(
    title = "Visualizing Funnel Geometry in Centred Parameterisation",
    subtitle = "Subject 11 shows the classic funnel shape collapsing to 0 at low group-level SD",
    x = "Subject Random Effect (u)",
    y = "log(Group Standard Deviation)"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

---

## Step 4: Run `fibr` Diagnostics

### 4.1 Parameterisation Advisor

We pass the centred fit to `fibr::smoothbp_advisor()`. This function evaluates the relationship between the group-level random-effect SD ($\sigma_{re}$) and the individual deviations to identify where the funnel geometry is degrading the sampler:

```{r run-advisor}
# Diagnose the centred model
advice <- fibr::smoothbp_advisor(fit_centred)
print(advice)
```

Look at the advisor's recommendation for the second breakpoint (`omega2`):
* **Subjects 1 to 10**: Recommended for the **centred** parameterisation. Since they have a second knot, the location $\omega_2$ is well-identified.
* **Subjects 11 to 20**: Recommended for the **non-centred** parameterisation. Because they do not have a second knot, the location $\omega_2$ is unidentified and gets trapped in a funnel.

### 4.2 Holonomy Diagnostic and Eigenspectrum

To understand the geometric obstruction causing these issues, we can run `fibr::holonomy_diagnostic()` to measure the statistical fiber memory (holonomy) of the second change-point location (`omega2`). 

This scans the chain for base-space loops in $(\omega_2, \sigma_{re, \omega_2})$ and estimates the linear transport map $H$ representing how the fiber variables (the subject-specific random deviations `omega2_re_subject_j`) shift or contract when the base hyperparameters return to their starting points.

```{r run-holonomy-diagnostic, fig.width=6, fig.height=5, fig.cap="Eigenspectrum of the transport map H for omega2. The dominant eigenvalue near 0.96 indicates extremely strong fiber memory: when the base hyperparameters return near a starting point after a loop, the random effects retain ~96% of their original values, indicating a severe geometric obstruction in the centred parameterisation."}
base_vars <- c("omega2_(Intercept)", "sigma_re_omega2")
fiber_vars <- paste0("omega2_re_subject_", 1:20)

hd <- fibr::holonomy_diagnostic(
  chain             = fit_centred$draws,
  base_vars         = base_vars,
  fiber_vars        = fiber_vars,
  min_gap           = 3L,          # within the autocorrelation time
  n_bootstrap       = 100L,
  residualize_fiber = TRUE
)
print(hd)

# Plot the eigenspectrum
plot(hd)
```

**Interpreting the eigenspectrum:** The presence of a dominant eigenvalue very close to 1.0 (specifically $\lambda_1 \approx 0.96$) on the real axis confirms that the centred parameterisation exhibits strong, non-trivial holonomy. Because the group standard deviation and individual deviations are coupled, the sampler cannot move freely without being dragged back to its starting state, confirming that a reparameterisation is required.


---

## Step 5: Fit the Optimized Model (Full Non-Centring)

The fibr advisor diagnoses which subjects are in the funnel regime. When any groups are flagged, the recommended remedy is to refit with `reparameterise = "omega"`, which applies full non-centring across all omega random effects:

```{r fit-optimized, message=FALSE}
t0 <- Sys.time()
fit_opt <- smoothbp_ss(
  formula = y ~ tau,
  b0      = ~ 1,
  b1      = ~ 1,
  deltas  = list(~ 1 + (1|subject), ~ 0 + (1|subject)),
  omega   = list(~ 1 + (1|subject), ~ 1 + (1|subject)),
  rho     = list(~ 1, ~ 1),
  data    = dat,
  priors  = smoothbp_priors(
    omega = prior_normal(3.0, 1.5, lb = 0.0)
  ),
  spike   = prior_spike_slab(pi = 0.5),
  reparameterise = "omega",
  chains  = 2L,
  iter    = 2000L,
  warmup  = 1000L,
  seed    = 777L,
  .verbose = FALSE
)
t_opt <- difftime(Sys.time(), t0, units = "secs")
```

---

## Step 6: Traceplot Comparison

Let's compare the MCMC traceplots of the population-level second changepoint location `omega2_(Intercept)` before and after applying non-centring:

```{r plot-traceplots, fig.width=7, fig.height=5}
draws_opt <- as_draws_df(fit_opt$draws)

df_trace <- data.frame(
  Iteration = rep(1:1000, 2),
  Chain = factor(rep(rep(1:2, each = 1000), 2)),
  Value = c(
    draws_centred$`omega2_(Intercept)`[draws_centred$.chain == 1],
    draws_centred$`omega2_(Intercept)`[draws_centred$.chain == 2],
    draws_opt$`omega2_(Intercept)`[draws_opt$.chain == 1],
    draws_opt$`omega2_(Intercept)`[draws_opt$.chain == 2]
  ),
  Parameterisation = rep(c("Centred (Naive)", "Non-Centred (reparameterise = 'omega')"), each = 2000)
)

ggplot(df_trace, aes(x = Iteration, y = Value, color = Chain)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~Parameterisation, scales = "free_y", ncol = 1) +
  labs(
    title = "Traceplot Comparison: Population intercept omega2",
    subtitle = "Non-centring resolves sticky chains and improves mixing",
    x = "Post-warmup Iterations",
    y = "Parameter Value"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")
```

Under the naive centred parameterisation, the chains mix poorly. The non-centred parameterisation (`reparameterise = "omega"`) allows the sampler to escape the funnel's flat gradients, leading to healthy, well-mixed chains.

---

## Step 7: Evaluate Knot Detection

By resolving the MCMC geometry issues, the Gibbs indicator sampler can now cleanly separate the subjects. Let's calculate the Posterior Inclusion Probabilities (PIPs) for each subject's second change-point ($\delta_{2,s}$):

```{r evaluate-pips}
pips <- pip(fit_opt)

# Format results
res <- data.frame(
  subject = 1:20,
  true_knots = c(rep("2 Knots", 10), rep("1 Knot", 10)),
  pip_delta2 = sapply(1:20, function(s) {
    pips$pip[pips$parameter == paste0("delta2_re_subject_", s)]
  })
)

print(res)
```

Let's visualize the PIPs. We expect all subjects with 2 knots to have a PIP near 1, and all subjects with 1 knot to have a PIP near 0:

```{r plot-pips, fig.width=7, fig.height=4}
ggplot(res, aes(x = factor(subject), y = pip_delta2, fill = true_knots)) +
  geom_bar(stat = "identity", width = 0.7, alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(
    title = "Posterior Inclusion Probabilities (PIPs) for the Second Breakpoint",
    subtitle = "A PIP > 0.5 threshold perfectly classifies the two groups",
    x = "Participant ID", y = "PIP (delta2)", fill = "True Structure"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

---

## Step 8: Benchmarking against Stan (`brms`)

To benchmark performance, we fit the equivalent model using **Stan (via `brms`)**. 
Because Stan uses a standard Hamiltonian Monte Carlo (NUTS) sampler, it does not support discrete spike-and-slab indicator variables ($\gamma \in \{0,1\}$). Thus, we compare against a standard hierarchical model in `brms` (estimating the random effects directly under normal priors).

The `brms` model is fitted dynamically below:

```{r fit-brms, message=FALSE, results='hide'}
bf_mod <- brms::bf(
  y ~ b0 + b1 * (tau - omega1) +
        delta1 * (tau - omega1) / (1 + exp(-(tau - omega1) * rho1)) +
        delta2 * (tau - omega2) / (1 + exp(-(tau - omega2) * rho2)),
  b0 ~ 1,
  b1 ~ 1,
  delta1 ~ 1 + (1 | subject),
  delta2 ~ 0 + (1 | subject),
  omega1 ~ 1 + (1 | subject),
  omega2 ~ 1 + (1 | subject),
  rho1 ~ 1,
  rho2 ~ 1,
  nl = TRUE
)

priors_brms <- c(
  prior(normal(5, 2), nlpar = "b0"),
  prior(normal(-1, 1), nlpar = "b1"),
  prior(normal(2, 1), nlpar = "delta1"),
  prior_string("normal(3, 1.5)", nlpar = "omega1", lb = 0),
  prior_string("normal(3, 1.5)", nlpar = "omega2", lb = 0),
  prior(normal(4, 1), nlpar = "rho1", lb = 0),
  prior(normal(4, 1), nlpar = "rho2", lb = 0)
)

init_fun <- function() list(
  b_b0     = array(5.0),
  b_b1     = array(-1.0),
  b_delta1 = array(2.0),
  b_omega1 = array(2.0),
  b_omega2 = array(4.0),
  b_rho1   = array(4.0),
  b_rho2   = array(4.0)
)

t0 <- Sys.time()
fit_brms <- brms::brm(
  bf_mod,
  data    = dat,
  prior   = priors_brms,
  chains  = 2, iter = 1000, warmup = 500,
  seed    = 777, refresh = 0,
  init    = init_fun,
  control = list(adapt_delta = 0.85)
)
t_brms <- difftime(Sys.time(), t0, units = "secs")
```

---

## Sampler Performance Comparison

We assemble the comparison table dynamically from the fitted models, calculating the execution time, the number of post-warmup divergent transitions, the Bulk Effective Sample Size (ESS) for the population intercept of the second changepoint ($\omega_{2}$), and the sampling speed in ESS per second:

```{r performance-table, echo=FALSE, message=FALSE}
get_metrics <- function(fit, time_taken, name, is_brms = FALSE) {
  if (is_brms) {
    draws <- posterior::as_draws_df(fit)
    s <- posterior::summarise_draws(draws)
    var_row <- s[s$variable == "b_omega2_Intercept", ]
    divs <- sum(brms::nuts_params(fit)$value[brms::nuts_params(fit)$parameter == "divergent__"])
  } else {
    draws <- fit$draws
    s <- posterior::summarise_draws(draws)
    var_row <- s[s$variable == "omega2_(Intercept)", ]
    divs <- fit$n_divergent
  }
  
  ess <- var_row$ess_bulk
  rhat <- var_row$rhat
  ess_sec <- ess / as.numeric(time_taken)
  
  data.frame(
    Model = name,
    Time_Sec = round(as.numeric(time_taken), 1),
    Divergences = divs,
    Omega2_ESS = round(ess, 1),
    Omega2_Rhat = round(rhat, 3),
    ESS_per_Sec = round(ess_sec, 2),
    stringsAsFactors = FALSE
  )
}

m_centred <- get_metrics(fit_centred, t_centred, "smoothbp_ss (Centred)")
m_opt     <- get_metrics(fit_opt,     t_opt,     "smoothbp_ss (Full NC)")
m_brms    <- get_metrics(fit_brms,    t_brms,    "brms (Stan / NUTS)", is_brms = TRUE)

comparison <- rbind(m_centred, m_opt, m_brms)
knitr::kable(comparison, format = "markdown")
```

### Key Takeaways
1. **Stan MCMC Failure**: Stan (via `brms`) suffers heavily from the funnel geometry because it uses a centred parameterisation by default for non-linear random effects. With 10 subjects lacking a second knot, Stan gets completely stuck in the funnel, resulting in **divergent transitions**, an $\hat{R}$ of **> 1.1** (indicating failure to converge), and a Bulk ESS of just **~3.0** after ~2.5 minutes.
2. **Full NC Advantage**: By using `fibr`'s diagnostic to identify the funnel regime and refitting with `reparameterise = "omega"`, `smoothbp_ss` achieves healthy convergence ($\hat{R} \approx 1.02$) and a high Bulk ESS in only ~45 seconds.
3. **Efficiency Gain**: The optimized `smoothbp_ss` model with full non-centring achieves an efficiency that is **over 200 times faster** (in terms of ESS/sec) than Stan.
