---
title: "Getting Started with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`smoothbp` fits hierarchical piecewise regression models in which the
transition between two linear regimes is smoothed by a logistic sigmoid.
The observation-level mean function is:

$$
\mu_i = b0_i + b1_i(\tau_i - \omega_i) + b2_i(\tau_i - \omega_i)\,\sigma\!\left[(\tau_i - \omega_i)\rho_i\right]
$$

where

| Symbol | Role |
|--------|------|
| $\tau$ | time / covariate |
| $\omega$ | change-point location |
| $\rho$ | transition sharpness ($\rho \to \infty$ → hard kink) |
| $b0$ | level at the change-point |
| $b1$ | pre-change slope |
| $b1 + b2$ | post-change slope (asymptotically) |

Each of $b0$, $b1$, $b2$, $\omega$, $\rho$ has its own linear predictor
specified via a formula. $b0$ may also include a random intercept
`(1 | group)` for clustered / repeated-measures data.

Posterior inference uses a **Metropolis-within-Gibbs sampler written in
Rust** for speed:

- $b0$, $b1$, $b2$, random intercepts — exact Gibbs (conjugate normal,
  non-centred reparameterisation for random effects).
- $\omega$, $\rho$ — random-walk Metropolis with adaptive step sizes
  during warmup targeting ~23% acceptance.
- $\sigma$, $\sigma_u$ — exact Gibbs (inverse-gamma conjugate).

---

## Simulating data

`simulate_smoothbp()` generates synthetic data from the model.  It is
useful for testing and for understanding how the parameters relate to the
observable trajectory shape.

```{r simulate}
library(smoothbp)

dat <- simulate_smoothbp(
  n_subj    = 20,
  n_obs     = 8,
  b0        = 5.0,   # level at change-point
  b1        = -0.3,  # pre-change slope
  b2        =  1.2,  # change in slope
  omega     =  3.0,  # change-point location
  rho       =  4.0,  # transition sharpness
  sigma     =  0.4,  # residual SD
  sigma_u   =  0.5,  # between-subject SD
  tau_range = c(0, 6),
  seed      = 42
)

head(dat)
#>   subject       tau       mu        y
#> 1       1 0.0000000 6.588... 7.107...
#> ...

# Inspect the data-generating values
true_params(dat)
#> Data-generating parameters:
#>   b0:        5
#>   b1:        -0.3
#>   ...
```

The returned data frame carries the true values as an attribute:

```{r true-params-attr}
attr(dat, "true_params")
```

Unbalanced designs are supported by passing a vector to `n_obs`:

```{r unbalanced}
dat_unbal <- simulate_smoothbp(n_subj = 10, n_obs = c(4, 6, 8, 5, 7, 6, 8, 4, 5, 6))
```

---

## Fitting the model

Pass a two-sided formula identifying the response and time variable, then
one one-sided formula per parameter component.

```{r fit}
fit <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),   # random intercept per subject
  b1      = ~ 1,
  b2      = ~ 1,
  omega   = ~ 1,
  rho     = ~ 1,
  data    = dat,
  priors  = smoothbp_priors(
    omega = prior_normal(3, 2, lb = 0)
  ),
  chains  = 4L,
  iter    = 2000L,
  warmup  = 1000L,
  seed    = 42L
)
```

### Progress bar

While sampling, each chain prints an ASCII progress bar that distinguishes
warmup from sampling:

```
Chain 1/4
  Chain 1/4 [                    ]   0%  (warmup  )  iter 0/2000
  Chain 1/4 [~~                  ]  10%  (warmup  )  iter 200/2000
  ...
  Chain 1/4 [~~~~~~~~~~          ]  50%  (sampling)  iter 1000/2000
  Chain 1/4 [~~~~~~~~~~==        ]  60%  (sampling)  iter 1200/2000
  ...
  Chain 1/4 [~~~~~~~~~~==========] 100%  (done    )  iter 2000/2000
```

`~` marks the warmup portion; `=` marks post-warmup sampling.  Pass
`.verbose = FALSE` to suppress output.

### Parallel chains

By default chains run sequentially with the full progress bar.  Pass
`cores` to run them in parallel using Rayon:

```{r fit-parallel}
fit <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),
  data    = dat,
  priors  = smoothbp_priors(omega = prior_normal(3, 2, lb = 0)),
  chains  = 4L, iter = 2000L, warmup = 1000L, seed = 42L,
  cores   = 4L    # run all 4 chains concurrently
)
#> Running 4 chains in parallel on 4 thread(s)...
#> All 4 chains complete.
```

The per-iteration progress bar is suppressed in parallel mode (interleaved
output from multiple threads would be unreadable); only the start/done
messages are printed.  Results are **identical** to the sequential run for
the same `seed` — each chain's seed is derived deterministically.

To make parallel the default for a project, add this to `.Rprofile`:

```{r rprofile, eval=FALSE}
options(smoothbp.cores = parallel::detectCores())
```

### Posterior summary

`print()` displays results in labelled sections; `summary()` returns a
data frame.  The `effects` argument (accepted by both) filters which
parameters are shown — mirroring `brms`:

```{r summary}
print(fit)                               # all sections (default)
print(fit, effects = "fixed")            # population-level only
summary(fit, effects = "fixed")          # same, as a data frame
summary(fit, effects = c("fixed", "ran_pars"))  # fixed + sigma_u, no u[j]
summary(fit, effects = "ran_vals")       # individual u[j] only
```

| `effects` value | Parameters shown |
|-----------------|-----------------|
| `"fixed"` | Regression coefficients + $\sigma$ |
| `"ran_pars"` | Random-effect SD $\sigma_u$ |
| `"ran_vals"` | Individual deviations $u_j$ |
| `"all"` | Everything (default) |

---

## Diagnostics

### Trace plots with mixing indicators

`trace_plot()` annotates each panel with $\hat{R}$ and bulk-ESS, and
highlights poorly mixing parameters with a red background.

```{r trace}
trace_plot(fit)
```

Parameters with $\hat{R} > 1.05$ or bulk-ESS $< 100$ are flagged with a
⚠ symbol in the panel header, a red-tinted background, and a warning
message listing the affected parameters.  The thresholds are adjustable:

```{r trace-strict}
trace_plot(fit, rhat_thresh = 1.01, ess_thresh = 400)
```

`type = "density"` or `type = "both"` produces posterior density overlays
instead of (or alongside) the trace:

```{r density}
trace_plot(fit, type = "both")
```

`plot(fit)` is a convenient alias for `trace_plot(fit)`.

### Posterior predictive check

```{r pp-check}
pp_check(fit)
```

Overlays the marginal predictive densities from 50 posterior draws over
the observed response density.

---

## Parameter recovery

When fitting a model to data from `simulate_smoothbp()`, use
`recovery_plot()` to compare the posterior against the known truth.

```{r recovery}
recovery_plot(fit, dat)
```

Each row shows one population-level parameter:

- **Filled circle** — posterior mean
- **Horizontal line** — 95% credible interval (width set by `level`)
- **×** — true data-generating value
- **Green** — interval contains truth; **red** — interval misses truth

The subtitle reports the overall coverage count and turns red if any
interval misses the truth.

```{r recovery-90}
recovery_plot(fit, dat, level = 0.90)
```

---

## Prior specification

`smoothbp_priors()` accepts either a single prior applied to all
coefficients of a component, or a named list for per-coefficient overrides.
All priors are truncated normal (via `prior_normal()`) or inverse-gamma
(via `prior_invgamma()`).

```{r priors}
smoothbp_priors(
  b0    = prior_normal(0, 10),
  b1    = prior_normal(0, 5),
  b2    = prior_normal(0, 5),
  omega = list(
    "(Intercept)"    = prior_normal(3, 2, lb = 0, ub = 6),
    "groupTreatment" = prior_normal(0, 2)
  ),
  rho   = prior_normal(4, 2, lb = 0),
  sigma = prior_invgamma(shape = 2, scale = 1)
)
```

The `lb` and `ub` arguments to `prior_normal()` impose hard bounds.
`omega` should always have `lb = 0` (it is a location on the time axis).
`rho` should have `lb = 0` (sharpness must be positive).

---

## Hypothesis tests and evidence ratios

`hypothesis()` evaluates directional hypotheses and linear contrasts directly
against the posterior draws.

```{r hypothesis-directional}
# Is the change in slope positive?
hypothesis(fit, "b2_(Intercept) > 0")

# Does the change-point fall before time 4?
hypothesis(fit, "omega_(Intercept) < 4")
```

Multiple hypotheses can be tested in one call:

```{r hypothesis-multiple}
hypothesis(fit, c(
  "b2_(Intercept) > 0",
  "b1_(Intercept) < 0",
  "omega_(Intercept) < 4",
  "b2_(Intercept) - b1_(Intercept) > 0"
))
```

Passing an expression **without** a comparison operator returns the full
posterior summary of that derived quantity:

```{r hypothesis-contrast}
# Posterior distribution of the net slope change
hypothesis(fit, "b2_(Intercept) - b1_(Intercept)")
```

### How evidence ratios are computed

For a directional hypothesis $H: A > B$ the function:

1. Computes the contrast $c = A - B$ for every posterior draw.
2. Reports $P(H) = P(c > 0)$ and the **evidence ratio**
   $ER = P(H) \,/\, (1 - P(H))$.

For $H: A < B$ the contrast is reversed so $P(H) = P(B - A > 0)$.  The
reported `Estimate` and credible interval always describe the contrast
(direction-adjusted), making it easy to see both the magnitude and the
uncertainty of the effect alongside the probability.

| Stars | ER | $P(H)$ |
|-------|----|--------|
| `***` | > 99 | > 0.99 |
| `**` | > 19 | > 0.95 |
| `*`  | > 3  | > 0.75 |

Point-null tests ($H: \theta = 0$) are not supported because they require
the Savage-Dickey density ratio.  For equivalence testing use
`bayestestR::rope()` on `posterior_draws(fit)`.

---

## Extracting results

```{r extract}
# Full draws_array (posterior package format)
posterior_draws(fit)

# Fitted values with 95% CI per observation
fitted(fit)

# Fitted draws matrix (n_draws × n_obs) — for custom summaries
fitted(fit, summary = FALSE)
```
