---
title: "Getting Started with smoothbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with smoothbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`smoothbp` fits hierarchical piecewise regression models in which the
transition between two linear regimes is smoothed by a logistic sigmoid.
The observation-level mean function is:

$$
\mu_i = b0_i + b1_i(\tau_i - \omega_i) + b2_i(\tau_i - \omega_i)\,\sigma\!\left[(\tau_i - \omega_i)\rho_i\right]
$$

where

| Symbol | Role |
|--------|------|
| $\tau$ | time / covariate |
| $\omega$ | change-point location |
| $\rho$ | transition sharpness ($\rho \to \infty$ → hard kink) |
| $b0$ | level at the change-point |
| $b1$ | pre-change slope |
| $b1 + b2$ | post-change slope (asymptotically) |

Each of $b0$, $b1$, $b2$, $\omega$, $\rho$ has its own linear predictor
specified via a formula. $b0$ may also include a random intercept
`(1 | group)` for clustered / repeated-measures data.

Posterior inference uses a **Metropolis-within-Gibbs sampler written in
Rust** for speed:

- $b0$, $b1$, $b2$, random intercepts — exact Gibbs (conjugate normal,
  non-centred reparameterisation for random effects).
- $\omega$, $\rho$ — random-walk Metropolis with adaptive step sizes
  during warmup targeting ~23% acceptance.
- $\sigma$, $\sigma_u$ — exact Gibbs (inverse-gamma conjugate).

---

## Simulating data

`simulate_smoothbp()` generates synthetic data from the model.  It is
useful for testing and for understanding how the parameters relate to the
observable trajectory shape.

```{r simulate}
library(smoothbp)

dat <- simulate_smoothbp(
  n_subj    = 20,
  n_obs     = 8,
  b0        = 5.0,   # level at change-point
  b1        = -0.3,  # pre-change slope
  b2        =  1.2,  # change in slope
  omega     =  3.0,  # change-point location
  rho       =  4.0,  # transition sharpness
  sigma     =  0.4,  # residual SD
  sigma_u   =  0.5,  # between-subject SD
  tau_range = c(0, 6),
  seed      = 42
)

head(dat)
#>   subject       tau       mu        y
#> 1       1 0.0000000 6.588... 7.107...
#> ...

# Inspect the data-generating values
true_params(dat)
#> Data-generating parameters:
#>   b0:        5
#>   b1:        -0.3
#>   ...
```

The returned data frame carries the true values as an attribute:

```{r true-params-attr}
attr(dat, "true_params")
```

Unbalanced designs are supported by passing a vector to `n_obs`:

```{r unbalanced}
dat_unbal <- simulate_smoothbp(n_subj = 10, n_obs = c(4, 6, 8, 5, 7, 6, 8, 4, 5, 6))
```

---

## Fitting the model

Pass a two-sided formula identifying the response and time variable, then
one one-sided formula per parameter component.

```{r fit}
fit <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),   # random intercept per subject
  b1      = ~ 1,
  b2      = ~ 1,
  omega   = ~ 1,
  rho     = ~ 1,
  data    = dat,
  priors  = smoothbp_priors(
    omega = prior_normal(3, 2, lb = 0)
  ),
  chains  = 4L,
  iter    = 2000L,
  warmup  = 1000L,
  seed    = 42L
)
```

### Progress bar

While sampling, each chain prints an ASCII progress bar that distinguishes
warmup from sampling:

```
Chain 1/4
  Chain 1/4 [                    ]   0%  (warmup  )  iter 0/2000
  Chain 1/4 [~~                  ]  10%  (warmup  )  iter 200/2000
  ...
  Chain 1/4 [~~~~~~~~~~          ]  50%  (sampling)  iter 1000/2000
  Chain 1/4 [~~~~~~~~~~==        ]  60%  (sampling)  iter 1200/2000
  ...
  Chain 1/4 [~~~~~~~~~~==========] 100%  (done    )  iter 2000/2000
```

`~` marks the warmup portion; `=` marks post-warmup sampling.  Pass
`.verbose = FALSE` to suppress output.

### Parallel chains

By default chains run sequentially with the full progress bar.  Pass
`cores` to run them in parallel using Rayon:

```{r fit-parallel}
fit <- smoothbp(
  formula = y ~ tau,
  b0      = ~ 1 + (1 | subject),
  data    = dat,
  priors  = smoothbp_priors(omega = prior_normal(3, 2, lb = 0)),
  chains  = 4L, iter = 2000L, warmup = 1000L, seed = 42L,
  cores   = 4L    # run all 4 chains concurrently
)
#> Running 4 chains in parallel on 4 thread(s)...
#> All 4 chains complete.
```

The per-iteration progress bar is suppressed in parallel mode (interleaved
output from multiple threads would be unreadable); only the start/done
messages are printed.  Results are **identical** to the sequential run for
the same `seed` — each chain's seed is derived deterministically.

To make parallel the default for a project, add this to `.Rprofile`:

```{r rprofile, eval=FALSE}
options(smoothbp.cores = parallel::detectCores())
```

### Posterior summary

`print()` displays results in labelled sections; `summary()` returns a
data frame.  The `effects` argument (accepted by both) filters which
parameters are shown — mirroring `brms`:

```{r summary}
print(fit)                               # all sections (default)
print(fit, effects = "fixed")            # population-level only
summary(fit, effects = "fixed")          # same, as a data frame
summary(fit, effects = c("fixed", "ran_pars"))  # fixed + sigma_u, no u[j]
summary(fit, effects = "ran_vals")       # individual u[j] only
```

| `effects` value | Parameters shown |
|-----------------|-----------------|
| `"fixed"` | Regression coefficients + $\sigma$ |
| `"ran_pars"` | Random-effect SD $\sigma_u$ |
| `"ran_vals"` | Individual deviations $u_j$ |
| `"all"` | Everything (default) |

---

## Diagnostics

### Trace plots with mixing indicators

`trace_plot()` annotates each panel with $\hat{R}$ and bulk-ESS, and
highlights poorly mixing parameters with a red background.

```{r trace}
trace_plot(fit)
```

Parameters with $\hat{R} > 1.05$ or bulk-ESS $< 100$ are flagged with a
⚠ symbol in the panel header, a red-tinted background, and a warning
message listing the affected parameters.  The thresholds are adjustable:

```{r trace-strict}
trace_plot(fit, rhat_thresh = 1.01, ess_thresh = 400)
```

`type = "density"` or `type = "both"` produces posterior density overlays
instead of (or alongside) the trace:

```{r density}
trace_plot(fit, type = "both")
```

`plot(fit)` is a convenient alias for `trace_plot(fit)`.

### Posterior predictive check

```{r pp-check}
pp_check(fit)
```

Overlays the marginal predictive densities from 50 posterior draws over
the observed response density.

---

## Parameter recovery

When fitting a model to data from `simulate_smoothbp()`, use
`recovery_plot()` to compare the posterior against the known truth.

```{r recovery}
recovery_plot(fit, dat)
```

Each row shows one population-level parameter:

- **Filled circle** — posterior mean
- **Horizontal line** — 95% credible interval (width set by `level`)
- **×** — true data-generating value
- **Green** — interval contains truth; **red** — interval misses truth

The subtitle reports the overall coverage count and turns red if any
interval misses the truth.

```{r recovery-90}
recovery_plot(fit, dat, level = 0.90)
```

---

## Prior specification

`smoothbp_priors()` accepts either a single prior applied to all
coefficients of a component, or a named list for per-coefficient overrides.
All priors are truncated normal (via `prior_normal()`) or inverse-gamma
(via `prior_invgamma()`).

```{r priors}
smoothbp_priors(
  b0    = prior_normal(0, 10),
  b1    = prior_normal(0, 5),
  b2    = prior_normal(0, 5),
  omega = list(
    "(Intercept)"    = prior_normal(3, 2, lb = 0, ub = 6),
    "groupTreatment" = prior_normal(0, 2)
  ),
  rho   = prior_normal(4, 2, lb = 0),
  sigma = prior_invgamma(shape = 2, scale = 1)
)
```

The `lb` and `ub` arguments to `prior_normal()` impose hard bounds.
`omega` should always have `lb = 0` (it is a location on the time axis).
`rho` should have `lb = 0` (sharpness must be positive).

### Spike-and-slab variable selection

When you have multiple covariates in the `b2` formula and want to identify
which ones genuinely contribute to the slope change, `smoothbp_ss()` extends
the model with spike-and-slab priors on the `b2` coefficients.  Covariates
excluded by the spike are structurally zeroed — both their `b2` effect and
the corresponding `omega`/`rho` terms are set to exactly zero.

See `vignette("spike-and-slab")` for a full tutorial with worked examples.

---

## Hypothesis tests and evidence ratios

`hypothesis()` evaluates directional hypotheses and linear contrasts directly
against the posterior draws.

```{r hypothesis-directional}
# Is the change in slope positive?
hypothesis(fit, "b2_(Intercept) > 0")

# Does the change-point fall before time 4?
hypothesis(fit, "omega_(Intercept) < 4")
```

Multiple hypotheses can be tested in one call:

```{r hypothesis-multiple}
hypothesis(fit, c(
  "b2_(Intercept) > 0",
  "b1_(Intercept) < 0",
  "omega_(Intercept) < 4",
  "b2_(Intercept) - b1_(Intercept) > 0"
))
```

Passing an expression **without** a comparison operator returns the full
posterior summary of that derived quantity:

```{r hypothesis-contrast}
# Posterior distribution of the net slope change
hypothesis(fit, "b2_(Intercept) - b1_(Intercept)")
```

### How evidence ratios are computed

For a directional hypothesis $H: A > B$ the function:

1. Computes the contrast $c = A - B$ for every posterior draw.
2. Reports $P(H) = P(c > 0)$ and the **evidence ratio**
   $ER = P(H) \,/\, (1 - P(H))$.

For $H: A < B$ the contrast is reversed so $P(H) = P(B - A > 0)$.  The
reported `Estimate` and credible interval always describe the contrast
(direction-adjusted), making it easy to see both the magnitude and the
uncertainty of the effect alongside the probability.

| Stars | ER | $P(H)$ |
|-------|----|--------|
| `***` | > 99 | > 0.99 |
| `**` | > 19 | > 0.95 |
| `*`  | > 3  | > 0.75 |

Point-null tests ($H: \theta = 0$) are not supported because they require
the Savage-Dickey density ratio.  For equivalence testing use
`bayestestR::rope()` on `posterior_draws(fit)`.

---

## Model comparison

### Leave-one-out cross-validation (LOO-IC)

Compare models using leave-one-out cross-validation with Pareto smoothed
importance sampling:

```{r loo}
# Compute LOO-IC for a single model
loo1 <- loo(fit)

# Fit an alternative model and compare
fit2 <- smoothbp(
  formula = y ~ tau,
  b0      = ~ Group + (1 | subject),  # different formula
  b1      = ~ 1,
  b2      = ~ 1,
  omega   = ~ 1,
  rho     = ~ 1,
  data    = dat,
  chains  = 4L, iter = 2000L, warmup = 1000L, seed = 42L
)

loo2 <- loo(fit2)

# Compare: negative elpd_diff favours the first model
loo::loo_compare(loo1, loo2)
#>       elpd_diff se_diff
#> fit1    0.0      0.0
#> fit2  -45.2     10.8  # model 1 is preferred
```

Negative `elpd_diff` indicates that the first model has better expected
log predictive density (higher is better).

### Watanabe-Akaike Information Criterion (WAIC)

Alternatively, use WAIC for model comparison:

```{r waic}
waic1 <- waic(fit)
waic2 <- waic(fit2)

loo::loo_compare(waic1, waic2)
```

### Comparing smoothbp and brms models

Because both `loo.smoothbp_fit` and `loo.brmsfit` return `psis_loo` objects,
`loo::loo_compare()` works directly across packages:

```{r loo-cross-package}
library(brms)

fit_linear <- brm(
  y ~ tau + (1 | subject),
  data      = dat,
  chains    = 4L, iter = 2000L, warmup = 1000L, seed = 42L,
  save_pars = save_pars(all = TRUE)  # required for bridge sampling
)

# LOO comparison: smoothbp (piecewise) vs brms (linear)
loo::loo_compare(loo(fit), loo(fit_linear))
#>            elpd_diff se_diff
#> model1       0.0       0.0
#> fit_linear -22.9       4.6   # piecewise model preferred
```

A negative `elpd_diff` for `fit_linear` indicates the smoothbp model has
better expected log predictive density.

### Bayes factors via bridge sampling

LOO quantifies **predictive** performance.  For **posterior model
probabilities** (Bayes factors), use bridge sampling.  The Bayes factor
BF10 = p(data | model 1) / p(data | model 2) penalises model complexity
automatically through the prior (Occam's razor), and so can be more
decisive than LOO when models differ substantially in their number of
parameters.

#### Computing the Bayes factor

The package registers a `bridge_sampler.smoothbp_fit` method that integrates
with the [bridgesampling](https://cran.r-project.org/package=bridgesampling)
package:

```{r bridge-sampler}
library(bridgesampling)

# Compute log marginal likelihoods
bs_piece  <- bridge_sampler(fit)          # smoothbp model
bs_linear <- bridge_sampler(fit_linear)   # brms model (save_pars required)

# Bayes factor: piecewise (numerator) vs linear (denominator)
bayes_factor(bs_piece, bs_linear)
#> Estimated Bayes factor in favor of x1 over x2: 2699738
```

Alternatively, use the convenience wrapper that handles both types:

```{r bayes-factor-wrapper}
# Dispatches on the first argument; smoothbp model must be first
bayes_factor(fit, fit_linear)
```

#### Assessing reliability with error_measures()

Bridge sampling is a stochastic approximation.  Always check the
uncertainty of the log marginal likelihood estimate before interpreting
the Bayes factor:

```{r error-measures}
error_measures(bs_piece)
#> $re2        # relative squared error
#> [1] 0.011
#>
#> $cv         # coefficient of variation (ideal: < 0.05)
#> [1] 0.106
#>
#> $percentage # expressed as a percentage
#> [1] "11%"
```

| `cv` value | Interpretation |
|---|---|
| < 5% | Reliable estimate |
| 5–15% | Moderate uncertainty; increase iterations if possible |
| > 15% | Unreliable; consider more iterations or use LOO instead |

A high CV most commonly means the MCMC chains did not fully explore the
posterior — check `trace_plot()` and increase `iter` before trusting the
Bayes factor.

#### When the change-point is weakly identified

When data are close to linear, the change-point parameters ($\omega$, $\rho$)
become weakly identified, chains mix poorly, and bridge sampling estimates
carry high uncertainty.  In this situation, LOO-IC is a more robust
comparison tool because it only requires the pointwise log-likelihood:

```{r loo-vs-bf}
# Check mixing before computing Bayes factors
trace_plot(fit, params = c("b2_(Intercept)", "omega_(Intercept)"))

# If Rhat > 1.05 or bulk-ESS < 400, prefer LOO over Bayes factors
loo::loo_compare(loo(fit), loo(fit_linear))
```

#### Two smoothbp models

The same workflow applies when both models are `smoothbp_fit` objects:

```{r bf-two-smoothbp}
fit_grouped <- smoothbp(
  y ~ tau,
  b0     = ~ group + (1 | subject),   # group effect on intercept
  data   = dat,
  chains = 4L, iter = 2000L, warmup = 1000L, seed = 43L
)

# Compare: random-effects-only vs group-effect model
bs_base    <- bridge_sampler(fit)
bs_grouped <- bridge_sampler(fit_grouped)

bayes_factor(bs_base, bs_grouped)
error_measures(bs_base)
error_measures(bs_grouped)
```

---

## Extracting results

### Posterior draws

Access the full posterior draws in `posterior` package format:

```{r extract-draws}
# Full draws_array (posterior package format)
posterior_draws(fit)
```

### Fitted values

Compute posterior distributions of the latent mean $\mu_i$ for each observation:

```{r fitted-training}
# Fitted values with 95% CI at training observations
fitted(fit)
#>   .observation fitted_mean fitted_Q2.5 fitted_Q97.5
#> 1            1      6.420      6.120       6.720
#> 2            2      ...
```

Access the full posterior draws instead of summaries:

```{r fitted-draws}
# Fitted draws matrix (n_draws × n_obs) — for custom summaries
draws_mat <- fitted(fit, summary = FALSE)

# Derive custom summaries, e.g. P(μ > 6) per observation
colMeans(draws_mat > 6)
```

### Marginal predictions

To obtain **population-level predictions** at arbitrary covariate values, 
provide `newdata` **without** the random-effect grouping variable:

```{r fitted-marginal}
# New time points, no subject specified → marginal (population-level)
newdata_marginal <- data.frame(
  tau = seq(0, 6, by = 0.5),
  # omit 'subject' to get population-level predictions
)

fitted(fit, newdata = newdata_marginal)
#>   .observation fitted_mean fitted_Q2.5 fitted_Q97.5
#> 1            1      6.310      6.010       6.610
#> ...
```

### Conditional predictions

To obtain **subject-specific (group-level) predictions**, include the 
random-effect grouping variable in `newdata`:

```{r fitted-conditional}
# Include 'subject' → conditional predictions for that subject
newdata_conditional <- data.frame(
  tau     = seq(0, 6, by = 0.5),
  subject = "1"  # must be a level from the training data
)

fitted(fit, newdata = newdata_conditional)
#>   .observation fitted_mean fitted_Q2.5 fitted_Q97.5
#> 1            1      6.512      6.212       6.812
#> ...
```

Subjects not present in the training data will be treated as marginal (u = 0) 
with a warning.
