---
title: "Hierarchical change-points: random timing, structural differences, and geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical change-points: random timing, structural differences, and geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
library(smoothbp)
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Overview

Two distinct sources of sampling difficulty arise in hierarchical change-point
models:

| Source | Mechanism | Remedy |
|---|---|---|
| **Random timing** | Subject change-points vary around a population mean; funnel geometry in $(\sigma_{\mathrm{re}}, u_j)$ space when $\sigma_{\mathrm{re}}$ is small | `reparameterise = "omega"` (NC), diagnosed via fibr |
| **Structural differences** | Some subjects lack a breakpoint entirely; the unidentified location collapses to the prior, creating an extreme funnel | `reparameterise = "omega"` (NC); fibr `prior_frac` flags it |

This vignette works through both scenarios in order, using fibr diagnostics to
connect geometry to remedy.

---

## Scenario A: Random change-point timing

### Why random change-points?

Hierarchical piecewise models are common in longitudinal and clinical data: a
treatment effect kicks in, a developmental milestone occurs, a patient transitions
between disease states.  When the timing of that transition — the change-point
$\omega$ — varies meaningfully between subjects, fitting a single population
change-point inflates residual variance and biases slope estimates.

The standard Bayesian solution is a hierarchical prior:

$$\omega_j = \bar\omega + u_j, \quad u_j \sim \mathcal{N}(0, \sigma_{\mathrm{re}}^2)$$

The key quantity for sampling difficulty is `prior_frac` — the fraction of
information about each subject's change-point that comes from the prior rather
than the data:

$$\text{prior\_frac}_j = \frac{\sigma_{\mathrm{re}}^{-2}}
  {\sigma_{\mathrm{re}}^{-2} + \sigma^{-2}\sum_{i \in j}
  \!\left(\tfrac{\partial\mu_i}{\partial\omega_{j,i}}\right)^{\!2}}$$

- `prior_frac` → 1: prior dominates → funnel regime → NC recommended
- `prior_frac` → 0: data dominates → centred parameterisation is efficient

### Simulating heterogeneous change-points

```{r sim-A}
set.seed(42)
J      <- 15L
n_j    <- 12L
N      <- J * n_j

omega_bar_true <- 3.0
sigma_re_true  <- 0.9   # meaningful subject variation
delta_true     <- 1.8
rho_true       <- 2.5
sigma_y_true   <- 0.4

u_omega_true <- rnorm(J, 0, sigma_re_true)
subject_id   <- rep(seq_len(J), each = n_j)
tau_vals     <- rep(seq(0.5, 5.5, length.out = n_j), times = J)

omega_i <- omega_bar_true + u_omega_true[subject_id]
d_i     <- tau_vals - omega_i
y_i     <- delta_true * d_i * plogis(rho_true * d_i) + rnorm(N, 0, sigma_y_true)

dat_A <- data.frame(subject = factor(subject_id), tau = tau_vals, y = y_i)

cat(sprintf("N = %d  |  J = %d  |  true sigma_re = %.2f  |  omega range [%.1f, %.1f]\n",
  N, J, sigma_re_true,
  min(omega_bar_true + u_omega_true), max(omega_bar_true + u_omega_true)))
```

```{r plot-A, fig.cap = "Observed trajectories. Dashed vertical lines mark the true per-subject change-points."}
true_omega <- omega_bar_true + u_omega_true
ggplot(dat_A, aes(x = tau, y = y, group = subject, colour = subject)) +
  geom_line(alpha = 0.5, linewidth = 0.5) +
  geom_vline(
    data = data.frame(subject = factor(seq_len(J)), omega = true_omega),
    aes(xintercept = omega, colour = subject),
    linetype = "dashed", alpha = 0.6, linewidth = 0.4
  ) +
  scale_colour_viridis_d(guide = "none") +
  labs(title = "Scenario A — heterogeneous change-points",
       x = "tau", y = "y") +
  theme_minimal(base_size = 12)
```

### Centred parameterisation (default)

```{r fit-A-cent}
priors_A <- smoothbp_priors(
  omega       = prior_normal(3, 1.5, lb = 0),
  sigma_re_om = prior_invgamma(2, 1)
)

fit_A_cent <- smoothbp(
  y ~ tau,
  b0     = ~ 1 + (1 | subject),
  omega  = list(~ 1 + (1 | subject)),
  deltas = list(~ 1),
  data   = dat_A,
  priors = priors_A,
  chains = 4L, iter = 2000L, warmup = 1000L,
  reparameterise = "none",
  seed = 42L, .verbose = FALSE
)
cat(sprintf("Centred — divergences: %d\n", fit_A_cent$n_divergent))
```

```{r summ-A-cent}
s_cent <- summary(fit_A_cent, effects = "fixed")
re_cent <- summary(fit_A_cent, effects = "ran_pars")
key_vars <- c("omega1_(Intercept)", "delta1_(Intercept)", "rho1_(Intercept)", "sigma")
print(s_cent[s_cent$variable %in% key_vars,
             c("variable", "mean", "Q2.5", "Q97.5", "Rhat", "Bulk_ESS")],
      row.names = FALSE)
cat(sprintf("\nsigma_re_omega1: mean = %.3f  Rhat = %.3f  ESS = %d\n",
            re_cent$mean[re_cent$variable == "sigma_re_omega1"],
            re_cent$Rhat[re_cent$variable == "sigma_re_omega1"],
            as.integer(re_cent$Bulk_ESS[re_cent$variable == "sigma_re_omega1"])))
```

In this data-rich regime — many observations per subject and substantial
between-subject variation — the centred parameterisation converges cleanly.
Rhat and ESS should both be healthy.

### Diagnosing geometry with fibr

Even when the centred fit looks clean, it is worth running the fibr diagnostic —
particularly if your own data has fewer observations per subject or smaller
between-subject variation.

```{r fibr-A, eval = requireNamespace("fibr", quietly = TRUE)}
library(fibr)
advice_A <- smoothbp_advisor(fit_A_cent, n_draws = 200L)
print(advice_A)
```

```{r fibr-A-plot, eval = requireNamespace("fibr", quietly = TRUE), fig.cap = "Prior fraction per subject. Values below the blue threshold confirm centred is efficient in this regime."}
plot(advice_A)
```

In this data-rich example `prior_frac` is low for most subjects — the likelihood
dominates and centred is efficient.  If your data flags subjects above the red
threshold (e.g. fewer observations, noisier measurements, or tighter
between-subject variation), refit with `reparameterise = "omega"`.

### Full NC parameterisation

For reference — and to illustrate the syntax for when fibr does flag a problem:

```{r fit-A-nc}
fit_A_nc <- smoothbp(
  y ~ tau,
  b0     = ~ 1 + (1 | subject),
  omega  = list(~ 1 + (1 | subject)),
  deltas = list(~ 1),
  data   = dat_A,
  priors = priors_A,
  chains = 4L, iter = 2000L, warmup = 1000L,
  reparameterise = "omega",
  seed = 42L, .verbose = FALSE
)
cat(sprintf("Full NC — divergences: %d\n", fit_A_nc$n_divergent))
```

### Parameter recovery

```{r recovery-A, fig.cap = "Posterior medians and 90% credible intervals for per-subject change-points vs true values (NC fit)."}
omega_posts <- summarise_draws(
  subset_draws(fit_A_nc$draws,
               variable = paste0("omega1_re_subject_", seq_len(J))),
  "median", ~quantile(.x, c(0.05, 0.95))
)
omega_bar_post <- median(as.vector(as_draws_matrix(
  subset_draws(fit_A_nc$draws, variable = "omega1_(Intercept)"))))

recovery_df <- data.frame(
  subject     = seq_len(J),
  true_omega  = true_omega,
  post_median = omega_bar_post + omega_posts$median,
  post_q05    = omega_bar_post + omega_posts$`5%`,
  post_q95    = omega_bar_post + omega_posts$`95%`
)

ggplot(recovery_df, aes(x = true_omega)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = post_q05, ymax = post_q95),
                width = 0.04, colour = "#2E8B57", alpha = 0.7) +
  geom_point(aes(y = post_median), colour = "#2E8B57", size = 2.5) +
  labs(title = "Scenario A — change-point recovery",
       subtitle = "Points = posterior median; bars = 90% CI",
       x = "True omega_j", y = "Posterior estimate") +
  coord_equal() + theme_minimal(base_size = 12)
```

---

## Scenario B: Structural group differences

### The problem

A subtler but more severe funnel arises when subjects differ not just in *when*
a change-point occurs but in *whether* it occurs at all.  Consider a 2-breakpoint
model fitted to a mixed cohort:

- **Group A** (subjects 1–10): two genuine structural breaks — $\delta_2 \neq 0$,
  so $\omega_2$ is identified by the data.
- **Group B** (subjects 11–20): only one break — $\delta_2 = 0$, so $\omega_2$ is
  completely unidentified.  Its subject-level deviation $u_{\omega_2}$ collapses
  to zero as $\sigma_{\mathrm{re},\omega_2} \to 0$, creating a severe Neal funnel.

Under centred parameterisation, HMC cannot traverse this funnel efficiently:
chains get stuck, divergences accumulate, and the spike-and-slab indicator for
$\delta_2$ fails to classify subjects correctly.

### Simulating structurally heterogeneous data

```{r sim-B, message = FALSE, warning = FALSE}
# Group A: two knots
dat_B1 <- simulate_smoothbp(
  n_subj    = 10L, n_obs = 15L,
  b0 = 5.0, b1 = -1.0,
  delta = c(2.0, -2.0), omega = c(2.0, 4.0), rho = c(4.0, 4.0),
  sigma = 0.15, sigma_u = 0.0, tau_range = c(0.5, 5.5), seed = 123L
)
dat_B1$subject <- as.integer(as.character(dat_B1$subject))

# Group B: one knot (delta2 = 0)
dat_B2 <- simulate_smoothbp(
  n_subj    = 10L, n_obs = 15L,
  b0 = 5.0, b1 = -1.0,
  delta = c(2.0, 0.0), omega = c(2.0, 4.0), rho = c(4.0, 4.0),
  sigma = 0.15, sigma_u = 0.0, tau_range = c(0.5, 5.5), seed = 456L
)
dat_B2$subject <- as.integer(as.character(dat_B2$subject)) + 10L

dat_B <- rbind(dat_B1, dat_B2)
dat_B$subject <- factor(dat_B$subject)
```

```{r plot-B, fig.height = 4, fig.cap = "Subject 1 (2 knots) vs Subject 11 (1 knot). The second structural break is absent in Group B."}
example_dat <- dat_B |>
  filter(subject %in% c(1, 11)) |>
  mutate(Group = factor(ifelse(subject == 1,
    "2 Knots (Subject 1)", "1 Knot (Subject 11)")))

ggplot(example_dat, aes(x = tau, y = y, colour = Group)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_line(aes(y = mu), linewidth = 1) +
  labs(title = "Scenario B — structural differences between groups",
       subtitle = "Solid lines show the true underlying trajectories",
       x = "tau", y = "y") +
  theme_minimal() + scale_colour_brewer(palette = "Set1")
```

### Centred fit (naive)

```{r fit-B-cent, message = FALSE}
t0 <- Sys.time()
fit_B_cent <- smoothbp_ss(
  y ~ tau, b0 = ~ 1, b1 = ~ 1,
  deltas  = list(~ 1 + (1 | subject), ~ 0 + (1 | subject)),
  omega   = list(~ 1 + (1 | subject), ~ 1 + (1 | subject)),
  rho     = list(~ 1, ~ 1),
  data    = dat_B,
  priors  = smoothbp_priors(omega = prior_normal(3.0, 1.5, lb = 0.0)),
  spike   = prior_spike_slab(pi = 0.5),
  reparameterise = "none",
  chains = 2L, iter = 2000L, warmup = 1000L,
  seed = 777L, .verbose = FALSE
)
t_cent <- difftime(Sys.time(), t0, units = "secs")
cat(sprintf("Centred — divergences: %d\n", fit_B_cent$n_divergent))
```

### Visualising the funnel

The unidentified $u_{\omega_2}$ for Group B subjects collapses toward zero as
$\sigma_{\mathrm{re}}$ shrinks, forming the classic funnel shape:

```{r plot-funnel, fig.height = 4, fig.cap = "Subject 1's omega2 RE is data-identified (no funnel). Subject 11's collapses to 0 as log(sigma_re) decreases — Neal's funnel."}
draws_cent <- as_draws_df(fit_B_cent$draws)

df_funnel <- data.frame(
  log_sigma_re = log(draws_cent$sigma_re_omega2),
  u_subj_1     = draws_cent$omega2_re_subject_1,
  u_subj_11    = draws_cent$omega2_re_subject_11
) |>
  pivot_longer(c(u_subj_1, u_subj_11), names_to = "Subject", values_to = "u") |>
  mutate(Subject = factor(ifelse(Subject == "u_subj_1",
    "Subject 1 (2 Knots — Identified)",
    "Subject 11 (1 Knot — Unidentified)")))

ggplot(df_funnel, aes(x = u, y = log_sigma_re, colour = Subject)) +
  geom_point(alpha = 0.4, size = 1.2) +
  facet_wrap(~ Subject) +
  labs(title = "Funnel geometry in centred parameterisation",
       x = "Subject random effect (u)", y = "log(sigma_re)") +
  theme_minimal() + scale_colour_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

### fibr diagnostics

```{r fibr-B, eval = requireNamespace("fibr", quietly = TRUE)}
advice_B <- fibr::smoothbp_advisor(fit_B_cent)
print(advice_B)
```

The advisor flags subjects 11–20 (Group B) at breakpoint 2 with high `prior_frac`:
their second change-point location is unidentified and trapped in the funnel.
Subjects 1–10 (Group A) are fine — their second knot is data-identified.

### Full NC fit

```{r fit-B-nc, message = FALSE}
t0 <- Sys.time()
fit_B_nc <- smoothbp_ss(
  y ~ tau, b0 = ~ 1, b1 = ~ 1,
  deltas  = list(~ 1 + (1 | subject), ~ 0 + (1 | subject)),
  omega   = list(~ 1 + (1 | subject), ~ 1 + (1 | subject)),
  rho     = list(~ 1, ~ 1),
  data    = dat_B,
  priors  = smoothbp_priors(omega = prior_normal(3.0, 1.5, lb = 0.0)),
  spike   = prior_spike_slab(pi = 0.5),
  reparameterise = "omega",
  chains = 2L, iter = 2000L, warmup = 1000L,
  seed = 777L, .verbose = FALSE
)
t_nc <- difftime(Sys.time(), t0, units = "secs")
cat(sprintf("Full NC — divergences: %d\n", fit_B_nc$n_divergent))
```

### Traceplot comparison

```{r plot-traces, fig.height = 5, fig.cap = "Population intercept omega2 before and after NC reparameterisation. NC resolves the sticky chains caused by the funnel."}
draws_nc <- as_draws_df(fit_B_nc$draws)

df_trace <- data.frame(
  Iteration = rep(1:1000, 4),
  Chain = factor(rep(rep(1:2, each = 1000), 2)),
  Value = c(
    draws_cent$`omega2_(Intercept)`[draws_cent$.chain == 1],
    draws_cent$`omega2_(Intercept)`[draws_cent$.chain == 2],
    draws_nc$`omega2_(Intercept)`[draws_nc$.chain == 1],
    draws_nc$`omega2_(Intercept)`[draws_nc$.chain == 2]
  ),
  Parameterisation = rep(c("Centred", "Full NC (reparameterise = 'omega')"),
                         each = 2000)
)

ggplot(df_trace, aes(x = Iteration, y = Value, colour = Chain)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ Parameterisation, scales = "free_y", ncol = 1) +
  labs(title = "Traceplot — population intercept omega2",
       x = "Post-warmup iteration", y = "Parameter value") +
  theme_minimal() + scale_colour_brewer(palette = "Set2")
```

### Posterior inclusion probabilities

With the geometry resolved, the spike-and-slab indicator correctly classifies
subjects. Subjects 1–10 (two knots) should have PIP ≈ 1 for $\delta_2$;
subjects 11–20 (one knot) should have PIP ≈ 0.

```{r pips-B}
pips <- pip(fit_B_nc)
res <- data.frame(
  subject    = 1:20,
  true_knots = c(rep("2 Knots", 10), rep("1 Knot", 10)),
  pip_delta2 = sapply(1:20, function(s)
    pips$pip[pips$parameter == paste0("delta2_re_subject_", s)])
)
print(res)
```

```{r plot-pips, fig.height = 4, fig.cap = "PIP for the second breakpoint. A 0.5 threshold perfectly classifies the two groups."}
ggplot(res, aes(x = factor(subject), y = pip_delta2, fill = true_knots)) +
  geom_col(width = 0.7, alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  labs(title = "Posterior inclusion probabilities — second breakpoint (delta2)",
       x = "Subject", y = "PIP", fill = "True structure") +
  theme_minimal() + scale_fill_brewer(palette = "Set1")
```

---

## Stan benchmark

For reference, the same Scenario B model fitted in Stan via `brms`.  Stan's NUTS
uses a centred parameterisation for non-linear random effects by default, so it
encounters the same funnel without the geometry-aware per-subject adaptation that
smoothbp provides.

```{r fit-brms, message = FALSE, results = "hide", eval = requireNamespace("brms", quietly = TRUE) && requireNamespace("cmdstanr", quietly = TRUE)}
library(brms)

bf_mod <- brms::bf(
  y ~ b0 + b1 * (tau - omega1) +
        delta1 * (tau - omega1) / (1 + exp(-(tau - omega1) * rho1)) +
        delta2 * (tau - omega2) / (1 + exp(-(tau - omega2) * rho2)),
  b0 ~ 1, b1 ~ 1,
  delta1 ~ 1 + (1 | subject), delta2 ~ 0 + (1 | subject),
  omega1 ~ 1 + (1 | subject), omega2 ~ 1 + (1 | subject),
  rho1 ~ 1, rho2 ~ 1, nl = TRUE
)

priors_brms <- c(
  prior(normal(5, 2),    nlpar = "b0"),
  prior(normal(-1, 1),   nlpar = "b1"),
  prior(normal(2, 1),    nlpar = "delta1"),
  prior_string("normal(3, 1.5)", nlpar = "omega1", lb = 0),
  prior_string("normal(3, 1.5)", nlpar = "omega2", lb = 0),
  prior(normal(4, 1),    nlpar = "rho1", lb = 0),
  prior(normal(4, 1),    nlpar = "rho2", lb = 0)
)

init_fun <- function() list(
  b_b0 = array(5.0), b_b1 = array(-1.0), b_delta1 = array(2.0),
  b_omega1 = array(2.0), b_omega2 = array(4.0),
  b_rho1 = array(4.0), b_rho2 = array(4.0)
)

t0 <- Sys.time()
fit_brms <- brms::brm(
  bf_mod, data = dat_B, prior = priors_brms,
  chains = 2, iter = 1000, warmup = 500,
  seed = 777, refresh = 0, init = init_fun,
  control = list(adapt_delta = 0.85)
)
t_brms <- difftime(Sys.time(), t0, units = "secs")
```

```{r benchmark-table, echo = FALSE, message = FALSE, eval = requireNamespace("brms", quietly = TRUE) && requireNamespace("cmdstanr", quietly = TRUE)}
get_metrics <- function(fit, time_taken, name, is_brms = FALSE) {
  if (is_brms) {
    s <- posterior::summarise_draws(posterior::as_draws_df(fit))
    var_row <- s[s$variable == "b_omega2_Intercept", ]
    divs <- sum(brms::nuts_params(fit)$value[
      brms::nuts_params(fit)$parameter == "divergent__"])
  } else {
    s <- posterior::summarise_draws(fit$draws)
    var_row <- s[s$variable == "omega2_(Intercept)", ]
    divs <- fit$n_divergent
  }
  data.frame(
    Model       = name,
    Time_Sec    = round(as.numeric(time_taken), 1),
    Divergences = divs,
    Omega2_ESS  = round(var_row$ess_bulk, 1),
    Omega2_Rhat = round(var_row$rhat, 3),
    ESS_per_Sec = round(var_row$ess_bulk / as.numeric(time_taken), 2),
    stringsAsFactors = FALSE
  )
}

comparison <- rbind(
  get_metrics(fit_B_cent, t_cent, "smoothbp_ss (Centred)"),
  get_metrics(fit_B_nc,   t_nc,   "smoothbp_ss (Full NC)"),
  get_metrics(fit_brms,   t_brms, "brms / Stan", is_brms = TRUE)
)
knitr::kable(comparison, format = "markdown")
```

---

## Summary

| Scenario | Centred result | fibr signal | Remedy |
|---|---|---|---|
| A: Random timing, data-rich | Clean convergence | `prior_frac` low | None needed |
| A: Random timing, few obs / small $\sigma_{\mathrm{re}}$ | Funnel: high Rhat, low ESS | `prior_frac` high | `reparameterise = "omega"` |
| B: Structural differences | Severe funnel, divergences | `prior_frac` high in the unidentified groups | `reparameterise = "omega"` |

The recommended workflow:

1. Fit centred (`reparameterise = "none"`, the default).
2. Check `fit$n_divergent` and key Rhat/ESS values.
3. Run `fibr::smoothbp_advisor()` — if any subjects (at any breakpoint) show
   high `prior_frac`, refit with `reparameterise = "omega"`.
