---
title: "Hierarchical change-points: random slopes and change-point locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical change-points: random slopes and change-point locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
HAS_FIBR <- NOT_CRAN && requireNamespace("fibr", quietly = TRUE)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = NOT_CRAN,
  fig.width = 7,
  fig.height = 5
)
library(smoothbp)
library(posterior)
library(ggplot2)
```

## Why random change-points?

Hierarchical piecewise models are common in longitudinal and clinical data: a
treatment effect kicks in, a developmental milestone occurs, a patient
transitions between disease states.  The timing of that transition — the
change-point $\omega$ — often varies meaningfully between subjects.  Fitting a
single population change-point ignores this variation, inflates residual
variance, and gives biased estimates of slope change.

The standard Bayesian solution is a hierarchical prior:

$$
\omega_j = \bar\omega + u_j, \quad u_j \sim \mathcal{N}(0, \sigma_{\mathrm{re}}^2)
$$

This introduces a well-known sampling difficulty: the **funnel geometry** of
$(\sigma_{\mathrm{re}}, u_j)$ space.  When $\sigma_{\mathrm{re}}$ is small,
the posterior concentrates in a narrow neck that HMC must traverse with a
tiny step size, degrading mixing.

This vignette demonstrates three approaches to the funnel, in increasing
sophistication:

| Approach | When to use | Key argument |
|---|---|---|
| **Centred** (default) | Many obs per subject, large $\sigma_{\mathrm{re}}$ | `reparameterise = "none"` |
| **Full NC** | Any funnel regime flagged by fibr | `reparameterise = "omega"` |

---

## Simulating heterogeneous change-points

```{r simulate}
set.seed(42)
J      <- 15L          # subjects
n_j    <- 12L          # observations per subject
N      <- J * n_j

# True parameters
omega_bar_true <- 3.0
sigma_re_true  <- 0.9   # meaningful subject variation
delta_true     <- 1.8   # slope change at changepoint
rho_true       <- 2.5   # transition sharpness
sigma_y_true   <- 0.4

# Simulate
u_omega_true <- rnorm(J, 0, sigma_re_true)
subject_id   <- rep(seq_len(J), each = n_j)
tau_vals     <- rep(seq(0.5, 5.5, length.out = n_j), times = J)

omega_i <- omega_bar_true + u_omega_true[subject_id]
d_i     <- tau_vals - omega_i
mu_i    <- delta_true * d_i * plogis(rho_true * d_i)
y_i     <- mu_i + rnorm(N, 0, sigma_y_true)

dat <- data.frame(
  subject = factor(subject_id),
  tau     = tau_vals,
  y       = y_i
)

cat(sprintf(
  "N = %d  |  J = %d  |  true sigma_re = %.2f  |  omega range [%.1f, %.1f]\n",
  N, J, sigma_re_true,
  min(omega_bar_true + u_omega_true),
  max(omega_bar_true + u_omega_true)
))
```

```{r plot-data, fig.cap = "Observed trajectories. Dashed vertical lines mark the true per-subject change-points."}
true_omega <- omega_bar_true + u_omega_true
ggplot(dat, aes(x = tau, y = y, group = subject, colour = subject)) +
  geom_line(alpha = 0.5, linewidth = 0.5) +
  geom_vline(
    data = data.frame(subject = factor(seq_len(J)), omega = true_omega),
    aes(xintercept = omega, colour = subject),
    linetype = "dashed", alpha = 0.6, linewidth = 0.4
  ) +
  scale_colour_viridis_d(guide = "none") +
  labs(title = "Simulated data — heterogeneous change-points",
       x = "tau", y = "y") +
  theme_minimal(base_size = 12)
```

---

## Step 1 — Centred parameterisation (default)

We start with the default centred parameterisation.  The `(1 | subject)`
syntax in the `omega` formula specifies that each subject has its own
change-point location, distributed around a shared population mean.

```{r fit-cent}
priors_used <- smoothbp_priors(
  omega       = prior_normal(3, 1.5, lb = 0),
  sigma_re_om = prior_invgamma(2, 1)
)

fit_cent <- smoothbp(
  y ~ tau,
  b0     = ~ 1 + (1 | subject),
  omega  = list(~ 1 + (1 | subject)),
  deltas = list(~ 1),
  data   = dat,
  priors = priors_used,
  chains = 4L, iter = 2000L, warmup = 1000L,
  reparameterise = "none",   # centred (default)
  seed = 42L
)
cat(sprintf("Centred  — divergences: %d\n", fit_cent$n_divergent))
```

```{r cent-summary}
s_cent <- summary(fit_cent, effects = "fixed")
re_cent <- summary(fit_cent, effects = "ran_pars")
key_vars <- c("omega1_(Intercept)", "delta1_(Intercept)", "rho1_(Intercept)", "sigma")
print(s_cent[s_cent$variable %in% key_vars,
             c("variable", "mean", "Q2.5", "Q97.5", "Rhat", "Bulk_ESS")],
      row.names = FALSE)
cat(sprintf("\nsigma_re_omega1: mean = %.3f  Rhat = %.3f  ESS = %d\n",
            re_cent$mean[re_cent$variable == "sigma_re_omega1"],
            re_cent$Rhat[re_cent$variable == "sigma_re_omega1"],
            as.integer(re_cent$Bulk_ESS[re_cent$variable == "sigma_re_omega1"])))
```

In this data-rich regime — many observations per subject and substantial
between-subject variation — the centred parameterisation converges cleanly.
Rhat and ESS should both be healthy.

---

## Step 2 — Diagnose geometry with fibr

Even when the centred fit looks clean, it is worth running the **fibr**
diagnostic — particularly if your own data has fewer observations per subject
or smaller between-subject variation.  The **fibr** package (Bindoff, 2026)
computes the Fisher information decomposition at each posterior draw.  The key
diagnostic is `prior_frac`: the fraction of information about each subject's
change-point that comes from the prior rather than the data.

$$
\text{prior\_frac}_j = \frac{\sigma_{\mathrm{re}}^{-2}}
  {\sigma_{\mathrm{re}}^{-2} + \sigma^{-2}\sum_{i \in j}
  \!\left(\tfrac{\partial\mu_i}{\partial\omega_{j,i}}\right)^{\!2}}
$$

- `prior_frac` → 1: prior dominates → funnel regime → NC recommended
- `prior_frac` → 0: data dominates → centred is efficient

`delta_j = sqrt(prior_frac_j)` is the recommended per-group NC mixing
fraction (from the Gaussian partial-NC theory of Papaspiliopoulos et al. 2007).

```{r advisor, eval = HAS_FIBR}
library(fibr)

advice <- smoothbp_advisor(fit_cent, n_draws = 200L)
print(advice)
```

```{r advisor-plot, eval = HAS_FIBR, fig.cap = "Prior fraction per subject at breakpoint 1. Groups with prior_frac above the red threshold are in the funnel regime and benefit from non-centred parameterisation."}
plot(advice)
```

In this data-rich example `prior_frac` will be low for most subjects —
the likelihood dominates and the centred parameterisation is efficient.
In your own data, if subjects are flagged with high `prior_frac` (fewer
observations, noisier measurements, or tighter between-subject variation),
refit with `reparameterise = "omega"`.

---

## Step 3 — Full NC: non-centred for all groups

If `smoothbp_advisor()` flags high `prior_frac` in your data, apply NC
uniformly.  This replaces $u_j$ with $z_j = u_j / \sigma_{\mathrm{re}}$,
where $z_j \sim N(0,1)$ independently of $\sigma_{\mathrm{re}}$, removing
the funnel geometry from the sampler entirely.

```{r fit-nc}
fit_nc <- smoothbp(
  y ~ tau,
  b0     = ~ 1 + (1 | subject),
  omega  = list(~ 1 + (1 | subject)),
  deltas = list(~ 1),
  data   = dat,
  priors = priors_used,
  chains = 4L, iter = 2000L, warmup = 1000L,
  reparameterise = "omega",   # NC for all groups
  seed = 42L
)
cat(sprintf("Full NC  — divergences: %d\n", fit_nc$n_divergent))
```

---

## Comparing mixing across parameterisations

In this data-rich regime both parameterisations converge well.  The benefit
of NC becomes clear when `prior_frac` is high — try reducing `n_j` or
`sigma_re_true` in the simulation above to see the funnel regime.

```{r compare-mixing, fig.cap = "Bulk ESS for the population-level omega intercept and sigma_re. In data-rich regimes both parameterisations perform similarly."}
fmt_row <- function(fit, label) {
  s  <- summary(fit, effects = "fixed")
  re <- summary(fit, effects = "ran_pars")
  om <- s[s$variable == "omega1_(Intercept)", ]
  sr <- re[re$variable == "sigma_re_omega1", ]
  data.frame(
    parameterisation = label,
    parameter        = c("omega1_(Intercept)", "sigma_re_omega1"),
    Rhat             = c(om$Rhat, sr$Rhat),
    Bulk_ESS         = c(as.integer(om$Bulk_ESS), as.integer(sr$Bulk_ESS)),
    stringsAsFactors = FALSE
  )
}

rows <- list(fmt_row(fit_cent, "Centred"),
             fmt_row(fit_nc,   "Full NC"))
comp <- do.call(rbind, rows)
comp$parameterisation <- factor(comp$parameterisation,
  levels = c("Centred", "Full NC"))

print(comp, row.names = FALSE)
```

```{r compare-ess-plot, fig.cap = "ESS for all per-subject omega RE across the three parameterisations."}
ess_rows <- function(fit, label) {
  vars <- paste0("omega1_re_subject_", seq_len(J))
  s <- summarise_draws(subset_draws(fit$draws, variable = vars), "ess_bulk")
  data.frame(parameterisation = label, ESS = s$ess_bulk)
}

ess_rows_list <- list(ess_rows(fit_cent, "Centred"),
                      ess_rows(fit_nc,   "Full NC"))
ess_df <- do.call(rbind, ess_rows_list)
ess_df$parameterisation <- factor(ess_df$parameterisation,
  levels = c("Centred", "Full NC"))

pal <- c("Centred" = "#B22222",
         "Full NC" = "#2E8B57")

ggplot(ess_df, aes(x = parameterisation, y = ESS, fill = parameterisation)) +
  geom_boxplot(width = 0.45, alpha = 0.85) +
  geom_jitter(width = 0.12, alpha = 0.45, size = 1.5) +
  geom_hline(yintercept = 400, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = pal, guide = "none") +
  labs(title = "Bulk ESS for per-subject changepoints",
       subtitle = "Dashed line = recommended minimum (400)",
       x = NULL, y = "Bulk ESS") +
  theme_minimal(base_size = 12)
```

---

## Parameter recovery

```{r recovery, fig.cap = "Posterior medians and 90% credible intervals for per-subject changepoints vs true values (full NC fit). The geometry-aware sampler recovers individual changepoints accurately."}
# Use the full NC fit for the recovery plot
omega_posts <- summarise_draws(
  subset_draws(fit_nc$draws,
               variable = paste0("omega1_re_subject_", seq_len(J))),
  "median",
  ~quantile(.x, c(0.05, 0.95))
)

omega_bar_post <- median(
  as.vector(as_draws_matrix(
    subset_draws(fit_nc$draws, variable = "omega1_(Intercept)")
  ))
)

recovery_df <- data.frame(
  subject     = seq_len(J),
  true_omega  = true_omega,
  post_median = omega_bar_post + omega_posts$median,
  post_q05    = omega_bar_post + omega_posts$`5%`,
  post_q95    = omega_bar_post + omega_posts$`95%`
)

ggplot(recovery_df, aes(x = true_omega)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = post_q05, ymax = post_q95),
                width = 0.04, colour = "#2E8B57", alpha = 0.7) +
  geom_point(aes(y = post_median), colour = "#2E8B57", size = 2.5) +
  labs(title = "Changepoint recovery: true vs posterior (full NC)",
       subtitle = "Points = posterior median; bars = 90% CI",
       x = "True omega_j", y = "Posterior estimate of omega_j") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

```{r hyperparams}
s_hyper <- summarise_draws(
  subset_draws(fit_nc$draws,
               variable = c("omega1_(Intercept)", "sigma_re_omega1",
                            "delta1_(Intercept)", "rho1_(Intercept)")),
  "mean", "sd", ~quantile(.x, c(0.05, 0.95))
)
truth_row <- c(omega_bar_true, sigma_re_true, delta_true, rho_true)
s_hyper$truth <- truth_row
print(s_hyper[, c("variable", "mean", "sd", "5%", "95%", "truth")])
```

---

## Summary

Two independent sources of sampling difficulty arise in random-omega models,
and smoothbp addresses each separately:

| Problem | Cause | Solution |
|---|---|---|
| Within-subject parameter coupling | Off-diagonal Fisher metric between $u_\omega$, $u_\delta$, $u_{b1}$ | Joint per-subject NUTS with analytic mass matrix |
| Hyperparameter funnel | $(\sigma_{\mathrm{re}}, u_j)$ geometry as $\sigma_{\mathrm{re}} \to 0$ | NC reparameterisation, diagnosed via fibr `prior_frac` |

Two parameterisation choices:

| Parameterisation | Syntax | Best for |
|---|---|---|
| Centred | `reparameterise = "none"` | Data-rich regimes (many obs per subject) |
| Full NC | `reparameterise = "omega"` | Any funnel regime; use fibr to diagnose |

The recommended workflow: fit centred, diagnose with `fibr::smoothbp_advisor()`,
and refit with `reparameterise = "omega"` only if groups are flagged with high
`prior_frac`.  In data-rich regimes the centred parameterisation is typically
sufficient and NC will make little difference.  The funnel regime — few
observations per subject, small $\sigma_{\mathrm{re}}$, or noisy measurements
— is where NC earns its keep.
