---
title: "Hierarchical change-points: random slopes and change-point locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical change-points: random slopes and change-point locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = NOT_CRAN,
  fig.width = 7,
  fig.height = 5
)
library(smoothbp)
library(posterior)
library(ggplot2)
```

## Why random change-points?

Hierarchical piecewise models are common in longitudinal and clinical data: a
treatment effect kicks in, a developmental milestone occurs, a patient
transitions between disease states.  The timing of that transition — the
change-point $\omega$ — often varies meaningfully between subjects.  Fitting a
single population change-point ignores this variation, inflates residual
variance, and gives biased estimates of slope change.

The standard Bayesian solution is a hierarchical prior:

$$
\omega_j = \bar\omega + u_j, \quad u_j \sim \mathcal{N}(0, \sigma_{\mathrm{re}}^2)
$$

The sampling challenge is well-known: the centred parameterisation creates a
**funnel geometry** in $(\sigma_{\mathrm{re}}, u_j)$ space.  When
$\sigma_{\mathrm{re}}$ is small the $u_j$ are constrained near zero and the
sampler must navigate a narrow neck, producing divergences and low effective
sample size.

`smoothbp` resolves this with a **joint per-subject NUTS step** that uses the
analytic within-subject Fisher metric as its mass matrix, capturing the
coupling between $u_j$, the slope change $\delta_j$, and the changepoint
location simultaneously.

---

## Simulating heterogeneous change-points

```{r simulate}
set.seed(42)
J      <- 15L          # subjects
n_j    <- 12L          # observations per subject
N      <- J * n_j

# True parameters
omega_bar_true <- 3.0
sigma_re_true  <- 0.9   # meaningful subject variation
delta_true     <- 1.8   # slope change at changepoint
rho_true       <- 2.5   # transition sharpness
sigma_y_true   <- 0.4

# Simulate
u_omega_true <- rnorm(J, 0, sigma_re_true)
subject_id   <- rep(seq_len(J), each = n_j)
tau_vals     <- rep(seq(0.5, 5.5, length.out = n_j), times = J)

omega_i <- omega_bar_true + u_omega_true[subject_id]
d_i     <- tau_vals - omega_i
mu_i    <- delta_true * d_i * plogis(rho_true * d_i)
y_i     <- mu_i + rnorm(N, 0, sigma_y_true)

dat <- data.frame(
  subject = factor(subject_id),
  tau     = tau_vals,
  y       = y_i
)

cat(sprintf(
  "N = %d  |  J = %d  |  true sigma_re = %.2f  |  omega range [%.1f, %.1f]\n",
  N, J, sigma_re_true,
  min(omega_bar_true + u_omega_true),
  max(omega_bar_true + u_omega_true)
))
```

```{r plot-data, fig.cap = "Observed trajectories. Subjects transition at different times (vertical lines show true per-subject changepoints)."}
true_omega <- omega_bar_true + u_omega_true
ggplot(dat, aes(x = tau, y = y, group = subject, colour = subject)) +
  geom_line(alpha = 0.5, linewidth = 0.5) +
  geom_vline(
    data = data.frame(subject = factor(seq_len(J)), omega = true_omega),
    aes(xintercept = omega, colour = subject),
    linetype = "dashed", alpha = 0.6, linewidth = 0.4
  ) +
  scale_colour_viridis_d(guide = "none") +
  labs(title = "Simulated data — heterogeneous change-points",
       x = "tau", y = "y") +
  theme_minimal(base_size = 12)
```

---

> **Note on convergence.** The centred parameterisation
> (`reparameterise = "none"`, the default) can exhibit slow mixing on the
> population-level intercept `omega1_(Intercept)` when between-subject
> variability (`sigma_re_omega`) is large — a well-known property of
> centred hierarchical models.  In such cases, prefer
> `reparameterise = "omega"`, which uses a non-centred parameterisation
> for the per-subject deviations and typically yields substantially higher
> ESS for the variance components.  As with any hierarchical model,
> always inspect trace plots and Rhat values before reporting results.

---

## Fitting with smoothbp

The `(1 | subject)` syntax extends naturally to the change-point location:

```{r fit-smoothbp}
fit_re <- smoothbp(
  y ~ tau,
  b0 = ~ 1 + (1 | subject),
  omega  = list(~ 1 + (1 | subject)),
  deltas = list(~ 1),
  data   = dat,
  priors = smoothbp_priors(
    omega       = prior_normal(3, 1.5, lb = 0),
    sigma_re_om = prior_invgamma(2, 1)
  ),
  chains         = 4L, iter = 2000L, warmup = 1000L,
  reparameterise = "omega",   # NC for omega RE: removes the sigma_re funnel
  seed           = 42L
)
```

```{r smoothbp-summary}
summary(fit_re)
```

```{r smoothbp-diag}
cat(sprintf("Divergences: %d\n", sum(fit_re$n_divergent)))

key <- c("omega1_(Intercept)", "sigma_re_omega1",
         paste0("omega1_re_subject_", 1:J))
s_re <- summarise_draws(
  subset_draws(fit_re$draws, variable = key),
  "mean", "sd", "rhat", "ess_bulk"
)
print(s_re, n = Inf)
```

---

## Comparison with Stan (centred parameterisation)

We fit the same model in Stan using the standard centred form
$u_j \sim \mathcal{N}(0, \sigma_{\mathrm{re}})$.  This is the natural
parameterisation but creates a funnel in $(\sigma_{\mathrm{re}}, u_j)$ space.

```{r stan-model, eval = NOT_CRAN && requireNamespace("cmdstanr", quietly = TRUE)}
library(cmdstanr)

stan_code <- "
data {
  int<lower=0> N;
  int<lower=1> J;
  array[N] int<lower=1,upper=J> subject;
  vector[N] tau;
  vector[N] y;
}
parameters {
  real delta;
  real<lower=0> rho;
  real omega_bar;
  real<lower=0> sigma_re;
  real<lower=0> sigma_y;
  vector[J] u_omega;          // centred: funnel geometry
}
model {
  delta     ~ normal(0, 3);
  rho       ~ normal(3, 2);
  omega_bar ~ normal(3, 1.5);
  sigma_re  ~ inv_gamma(2, 1);
  sigma_y   ~ exponential(2);
  u_omega   ~ normal(0, sigma_re);   // centred prior
  {
    vector[N] omega_i = omega_bar + u_omega[subject];
    vector[N] d = tau - omega_i;
    vector[N] mu = delta * d .* inv_logit(rho * d);
    y ~ normal(mu, sigma_y);
  }
}
"
stan_file <- write_stan_file(stan_code)
mod_stan  <- cmdstan_model(stan_file)

fit_stan <- mod_stan$sample(
  data    = list(N = N, J = J, subject = subject_id,
                 tau = tau_vals, y = y_i),
  chains  = 4L, parallel_chains = 4L,
  iter_warmup = 1000L, iter_sampling = 1000L,
  seed    = 42L, refresh = 0L
)

cat("Stan sampler diagnostics:\n")
fit_stan$diagnostic_summary()
```

---

## Head-to-head comparison

```{r comparison, eval = NOT_CRAN && requireNamespace("cmdstanr", quietly = TRUE)}
draws_stan <- fit_stan$draws()

# Key parameters common to both fits
re_smoothbp <- summarise_draws(
  subset_draws(fit_re$draws, variable = c("omega1_(Intercept)", "sigma_re_omega1")),
  "rhat", "ess_bulk"
)
re_stan <- summarise_draws(
  subset_draws(draws_stan, variable = c("omega_bar", "sigma_re")),
  "rhat", "ess_bulk"
)

cat("─── smoothbp (joint NUTS, Fisher metric) ───────────────────────\n")
cat(sprintf("  Divergences      : %d\n", sum(fit_re$n_divergent)))
cat(sprintf("  ESS sigma_re_om  : %.0f\n",
            re_smoothbp$ess_bulk[re_smoothbp$variable == "sigma_re_omega1"]))
cat(sprintf("  Rhat sigma_re_om : %.3f\n",
            re_smoothbp$rhat[re_smoothbp$variable == "sigma_re_omega1"]))

cat("\n─── Stan (centred parameterisation) ────────────────────────────\n")
n_div_stan <- sum(fit_stan$diagnostic_summary()$num_divergent)
cat(sprintf("  Divergences      : %d\n", n_div_stan))
cat(sprintf("  ESS sigma_re     : %.0f\n",
            re_stan$ess_bulk[re_stan$variable == "sigma_re"]))
cat(sprintf("  Rhat sigma_re    : %.3f\n",
            re_stan$rhat[re_stan$variable == "sigma_re"]))
```

```{r comparison-plot, fig.cap = "ESS for the hyperparameter sigma_re and all subject-level changepoints. smoothbp's geometry-aware sampler maintains high ESS across the full parameter vector; Stan's centred parameterisation degrades sharply for sigma_re in the funnel region.", eval = NOT_CRAN && requireNamespace("cmdstanr", quietly = TRUE)}
# ESS for each subject's changepoint
ess_smoothbp <- summarise_draws(
  subset_draws(fit_re$draws,
               variable = paste0("omega1_re_subject_", seq_len(J))),
  "ess_bulk"
) |> transform(model = "smoothbp", param = paste0("u_omega[", seq_len(J), "]"))

ess_stan <- summarise_draws(
  subset_draws(draws_stan,
               variable = paste0("u_omega[", seq_len(J), "]")),
  "ess_bulk"
) |> transform(model = "Stan (centred)", param = paste0("u_omega[", seq_len(J), "]"))

ess_compare <- rbind(
  data.frame(model = "smoothbp",         ess = ess_smoothbp$ess_bulk),
  data.frame(model = "Stan (centred)",   ess = ess_stan$ess_bulk)
)

ggplot(ess_compare, aes(x = model, y = ess, fill = model)) +
  geom_boxplot(width = 0.4, alpha = 0.8) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 400, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("smoothbp" = "#2E8B57", "Stan (centred)" = "#B22222"),
                    guide = "none") +
  labs(title = "ESS for subject-level changepoints (u_omega[j])",
       subtitle = "Dashed line = recommended minimum ESS (400)",
       x = NULL, y = "Bulk ESS") +
  theme_minimal(base_size = 12)
```

---

## Parameter recovery

```{r recovery, fig.cap = "Posterior medians and 90% credible intervals for subject-level changepoints vs true values. The joint NUTS step recovers individual changepoints accurately despite the hierarchical shrinkage."}
omega_posts <- summarise_draws(
  subset_draws(fit_re$draws,
               variable = paste0("omega1_re_subject_", seq_len(J))),
  "median",
  ~quantile(.x, c(0.05, 0.95))
)

# The subject RE is the deviation from omega_bar; add the intercept
omega_bar_post <- median(
  as.vector(as_draws_matrix(
    subset_draws(fit_re$draws, variable = "omega1_(Intercept)")
  ))
)

recovery_df <- data.frame(
  subject       = seq_len(J),
  true_omega    = true_omega,
  post_median   = omega_bar_post + omega_posts$median,
  post_q05      = omega_bar_post + omega_posts$`5%`,
  post_q95      = omega_bar_post + omega_posts$`95%`
)

ggplot(recovery_df, aes(x = true_omega)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = post_q05, ymax = post_q95),
                width = 0.04, colour = "steelblue", alpha = 0.7) +
  geom_point(aes(y = post_median), colour = "steelblue", size = 2.5) +
  labs(title = "Changepoint recovery: true vs posterior",
       subtitle = "Points = posterior median; bars = 90% CI",
       x = "True omega_j", y = "Posterior estimate of omega_j") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

```{r hyperparams}
# Hyperparameter recovery
s_hyper <- summarise_draws(
  subset_draws(fit_re$draws,
               variable = c("omega1_(Intercept)", "sigma_re_omega1",
                            "delta1_(Intercept)", "rho1_(Intercept)")),
  "mean", "sd", ~quantile(.x, c(0.05, 0.95))
)
truth_row <- c(omega_bar_true, sigma_re_true, delta_true, rho_true)
s_hyper$truth <- truth_row
print(s_hyper[, c("variable", "mean", "sd", "5%", "95%", "truth")])
```

---

## Summary

| | smoothbp (joint NUTS) | Stan (centred) |
|---|---|---|
| Parameterisation | Per-subject Fisher metric | Standard centred |
| Mass matrix | Within-subject G_s (analytic) | Diagonal Euclidean |
| Captures slope–changepoint coupling | Yes | No |
| Divergences | Low | Higher (funnel) |
| ESS for σ_re | High | Lower |
| Formula syntax | `omega = list(~ 1 + (1\|subject))` | Custom Stan code |

**Two sources of sampling difficulty — two solutions:**

| Problem | Source | Solution |
|---|---|---|
| Within-subject coupling | Off-diagonal Fisher metric between $u_\omega$, $u_\delta$, $u_{b1}$ | Joint per-subject NUTS with analytic mass matrix |
| Hyperparameter funnel | $(\sigma_\text{re}, u_j)$ coupling as $\sigma_\text{re} \to 0$ | `reparameterise = "omega"`: samples $z_j = u_j/\sigma_\text{re}$ with $N(0,1)$ prior |

`smoothbp` addresses both: `reparameterise = "omega"` removes the sigma funnel,
while the geometry-aware joint NUTS navigates the within-subject coupling.
Both are needed for robust inference with sparse data and wide priors.
