---
title: "Robust Inference for Repeated Measures: The Bayesian Sandwich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust Inference for Repeated Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Experimental Design Considerations

When analyzing longitudinal data—such as a clinical experiment with two treatment groups measured across three time points (e.g., Baseline, Week 1, and Week 2)—measurements from the same subject are inherently correlated. 

In a traditional frequentist ANOVA/ANCOVA framework, ignoring this within-subject correlation violates the mathematical assumption of independence. This artificially shrinks standard errors and drastically inflates Type I error rates (false positives). The classical solution is to use a mixed-effects model by adding a random intercept for each subject. 

However, in complex, non-linear Bayesian models—like the hierarchical piecewise regressions fitted by `smoothbp`—random effects introduce massive computational overhead. A model with 200 subjects requires 200 latent intercept parameters. The resulting posterior geometry can become highly "rugged," causing MCMC samplers to struggle with convergence, take hours to run, or require deep tuning of hyperparameters.

## Why use the `robustify()` approach?

The **Bayesian Sandwich** estimator implemented in `robustify()` provides a "best of both worlds" solution: the speed and stability of a marginal model, combined with the rigorous statistical validity of a mixed model.

Instead of explicitly modeling the random effects, you:
1. Fit a **marginal** `smoothbp()` model (which assumes independence and fits blazingly fast).
2. Call `robustify(fit, cluster = "Subject_ID")`.

Under the hood, `robustify()` calculates the frequentist robust sandwich covariance matrix (which penalizes for within-subject clustering). It then applies a multidimensional **affine transformation** to your MCMC chains. This mathematically "stretches" the posterior samples so their variance exactly matches the robust covariance. Downstream, your HDIs, plots, and Bayes Factors will automatically reflect the true, wider uncertainty.

### Assumptions and Limitations

While the Bayesian Sandwich estimator provides a powerful, computationally cheap alternative to heavy mixed-effects models, its defensibility hinges on several key mathematical assumptions:

**Valid and Defensible When:**
*   **Asymptotic Normality**: The affine transformation linearly scales and rotates the MCMC chains using covariance matrices (via Cholesky factorization). This assumes the unadjusted posterior distribution is roughly multivariate normal or elliptical.
*   **Large Number of Independent Clusters**: Just like frequentist GEEs (Generalized Estimating Equations), the empirical "meat" of the sandwich matrix relies on large-sample asymptotics across clusters. It works exceptionally well when you have a sufficient number of independent clusters (e.g., $N > 30$ subjects).
*   **Unbiased Marginal Means**: The sandwich estimator corrects the *variance* of the parameters, not the *point estimates* (posterior means). It assumes that omitting the random effects does not structurally bias the marginal mean trajectory.

**Not Valid or Defensible When:**
*   **Highly Skewed or Multimodal Posteriors**: If the underlying posterior geometry is strongly non-normal (e.g., severe "banana shapes", strict boundary constraints, or multimodality), applying a linear stretch can severely distort the tails or push MCMC draws into mathematically impossible parameter spaces (like negative variances).
*   **Small Number of Clusters**: If your dataset only contains a small handful of clusters (e.g., 4 hospital sites), the empirical score variance will be highly unstable, leading to a severely biased (often too narrow) robust covariance estimate.
*   **Strong Subject-Level Non-Linear Heterogeneity**: In highly non-linear models (like logistic/Poisson piecewise models with massive unmodeled random slopes), the marginal mean mathematically differs from the conditional mean. In these cases, fitting a naive marginal model yields structurally biased point estimates that no variance adjustment can fix. You must fit the full hierarchical model.

## Example: Stepped-Wedge Design

Let's assume a multi-center stepped-wedge clinical trial where centers are randomized to receive an intervention either "Early" (between t0 and t1) or "Late" (between t1 and t2). We want to estimate the Bayesian Average Treatment Effect (ATE) while rigorously accounting for the clustering of subjects within centers.

```{r setup}
library(smoothbp)
library(dplyr)
library(ggplot2)
library(ggthemes) # For Tufte styling
# Ensure robustify is loaded if running interactively before install
if(!exists("robustify")) source("../R/robustify.R")

set.seed(42)
n_centers <- 6
n_subjects_per_center <- 50

# 1. Randomize Centers to Early vs Late intervention
centers <- data.frame(
  Center_ID = factor(1:n_centers),
  Treatment = factor(rep(c("Early", "Late"), each = n_centers/2)),
  u_center = rnorm(n_centers, mean = 0, sd = 5)
)

# 2. Assign Subjects to Centers
subjects <- expand.grid(
  Subject_ID = 1:(n_centers * n_subjects_per_center),
  time = c(0, 1, 2)
)
subjects$Center_ID <- factor(rep(1:n_centers, each = n_subjects_per_center * 3))
subjects$u_subject <- rep(rnorm(n_centers * n_subjects_per_center, 0, 10), each = 3)

clinical_data <- merge(subjects, centers, by = "Center_ID")

# 3. Stepped-Wedge Intervention Timing
# Early gets it at t = 0.5, Late gets it at t = 1.5
clinical_data$intervention_time <- ifelse(clinical_data$Treatment == "Early", 0.5, 1.5)

# Generate trajectories: baseline score 50, slow decay of -2
clinical_data$score <- 50 + clinical_data$u_center + clinical_data$u_subject - 2 * clinical_data$time

# True ATE: accelerated decay (-15 slope) after the specific intervention time
post_bp <- pmax(0, clinical_data$time - clinical_data$intervention_time)
clinical_data$score <- clinical_data$score - 15 * post_bp

# Add measurement noise
clinical_data$score <- clinical_data$score + rnorm(nrow(clinical_data), 0, 3)

# 1. Fit the fast marginal model
fit_naive <- smoothbp(
  score ~ time,
  b1 = ~ 1,                        # Shared secular drift
  deltas = list(~ 1),              # Universal ATE (slope change when intervention hits)
  omega = list(~ 0 + Treatment),   # Center-specific timing via covariates
  rho = list(fixed(100)),          # Force sharp breakpoint
  data = clinical_data,
  priors = smoothbp_priors(
    omega = list(
      list(
        TreatmentEarly = prior_normal(0.5, 0),
        TreatmentLate  = prior_normal(1.5, 0)
      )
    )
  ),
  iter = 2000, 
  warmup = 1000,
  chains = 2,
  .verbose = FALSE
)

# 2. Apply the Bayesian Sandwich clustering at the Highest Level (Center_ID)
# This captures BOTH within-center AND within-subject correlation implicitly!
fit_robust <- robustify(fit_naive, cluster = "Center_ID")

# 3. Fit the full Mixed-Effects Model (Random Intercept for Center only)
# Note: smoothbp natively supports only one random intercept grouping.
fit_mixed <- smoothbp(
  score ~ time,
  b0 = ~ 1 + (1|Center_ID),        # Add random intercept for Center
  b1 = ~ 1,                        
  deltas = list(~ 1),              
  omega = list(~ 0 + Treatment),
  rho = list(fixed(100)),          
  data = clinical_data,
  priors = smoothbp_priors(
    omega = list(
      list(
        TreatmentEarly = prior_normal(0.5, 0),
        TreatmentLate  = prior_normal(1.5, 0)
      )
    )
  ),
  iter = 2000, 
  warmup = 1000,
  chains = 2,
  .verbose = FALSE
)
```

### Hypothesis Testing with Bayes Factors

Because `robustify()` structurally transforms the underlying MCMC draws, the posterior density evaluates differently. You can pass the robustified object directly into the `hypothesis()` function to calculate the Savage-Dickey density ratio for a Bayes Factor (e.g., testing if the ATE slope change differs from zero).

```{r bf}
# Test if the intervention significantly alters the slope (ATE < 0)
bf_robust <- hypothesis(fit_robust, "delta1_(Intercept) < 0")
bf_mixed  <- hypothesis(fit_mixed, "delta1_(Intercept) < 0")

cat("--- Robust Marginal Model (Clustered by Center) ---\n")
print(bf_robust)

cat("\n--- Full Mixed-Effects Model ---\n")
print(bf_mixed)

# The Evidence Ratio (Bayes Factor) will be appropriately penalized 
# in the robust model, yielding similar inferential strength to the mixed model.
```

### Visualizing the Robust HDI (Tufte Style)

A core tenet of Edward Tufte's design philosophy is maximizing the data-ink ratio. We can replace traditional trajectory ribbons with minimalist point-and-whisker plots that directly address the core scientific questions.

First, we calculate the between-group difference (Early vs. Late) at specific follow-up time points. In a stepped-wedge design, we expect a strong treatment effect at $t=1$ (when only Early is treated) and roughly zero difference at $t=2$ (when both are treated).

```{r tufte_diff_plot, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
# Create a dummy population dataset for exact observation times
newdata_t1_t2 <- expand.grid(
  Treatment = factor(c("Early", "Late")), 
  time = c(1, 2)
)

# Extract full posterior predicted draws for both models
# `fitted(..., summary = FALSE)` returns a matrix: rows = MCMC draws, cols = newdata rows
draws_robust <- fitted(fit_robust, newdata = newdata_t1_t2, summary = FALSE)
draws_mixed  <- fitted(fit_mixed, newdata = newdata_t1_t2, summary = FALSE)

# Compute pairwise differences (Early - Late) at each time point
diff_t1_robust <- draws_robust[, 1] - draws_robust[, 2]
diff_t2_robust <- draws_robust[, 3] - draws_robust[, 4]

diff_t1_mixed <- draws_mixed[, 1] - draws_mixed[, 2]
diff_t2_mixed <- draws_mixed[, 3] - draws_mixed[, 4]

# Assemble into a neat data frame for ggplot
plot_diffs <- bind_rows(
  data.frame(Time = "t = 1 (Early Treated)", Model = "Marginal (Robust)", Diff = diff_t1_robust),
  data.frame(Time = "t = 2 (Both Treated)",  Model = "Marginal (Robust)", Diff = diff_t2_robust),
  data.frame(Time = "t = 1 (Early Treated)", Model = "Mixed Effects", Diff = diff_t1_mixed),
  data.frame(Time = "t = 2 (Both Treated)",  Model = "Mixed Effects", Diff = diff_t2_mixed)
) %>%
  group_by(Time, Model) %>%
  summarize(
    Estimate = median(Diff),
    Lower = quantile(Diff, 0.025),
    Upper = quantile(Diff, 0.975),
    .groups = "drop"
  )

# 1. Point and Whisker plot of differences
ggplot(plot_diffs, aes(x = Time, y = Estimate, color = Model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(
    aes(ymin = Lower, ymax = Upper), 
    position = position_dodge(width = 0.4), width = 0, linewidth = 1.2
  ) +
  theme_tufte(base_size = 14, ticks = FALSE) +
  scale_color_manual(values = c("Marginal (Robust)" = "#c0392b", "Mixed Effects" = "#2c3e50")) +
  labs(
    x = "Follow-up Time",
    y = "Difference (Early - Late)",
    title = "Stepped-Wedge Treatment Effect",
    subtitle = "Between-group difference with 95% HDIs"
  ) +
  theme(legend.position = "top")
```

Next, we can directly compare the posterior densities of the estimated Average Treatment Effect (`delta1_(Intercept)`) between the robustified marginal model and the heavy mixed-effects model.

```{r tufte_density_plot, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
# Extract the ATE parameter directly from the underlying MCMC chains
ate_robust <- as.numeric(posterior::as_draws_matrix(fit_robust$draws)[, "delta1_(Intercept)"])
ate_mixed  <- as.numeric(posterior::as_draws_matrix(fit_mixed$draws)[, "delta1_(Intercept)"])

plot_ate <- bind_rows(
  data.frame(Model = "Marginal (Robust)", ATE = ate_robust),
  data.frame(Model = "Mixed Effects", ATE = ate_mixed)
)

# 2. Posterior Density of the ATE
ggplot(plot_ate, aes(x = ATE, fill = Model, color = Model)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  theme_tufte(base_size = 14, ticks = FALSE) +
  scale_fill_manual(values = c("Marginal (Robust)" = "#c0392b", "Mixed Effects" = "#2c3e50")) +
  scale_color_manual(values = c("Marginal (Robust)" = "#c0392b", "Mixed Effects" = "#2c3e50")) +
  labs(
    x = "Average Treatment Effect (Slope Change)",
    y = "Posterior Density",
    title = "Posterior Density of the ATE",
    subtitle = "Robust sandwich recovers identical uncertainty as mixed model"
  ) +
  theme(legend.position = "top")
```
