---
title: "Robust Inference for Repeated Measures: The Bayesian Sandwich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust Inference for Repeated Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Experimental Design Considerations

When analyzing longitudinal data—such as a clinical experiment with two treatment groups measured across three time points (e.g., Baseline, Week 1, and Week 2)—measurements from the same subject are inherently correlated. 

In a traditional frequentist ANOVA/ANCOVA framework, ignoring this within-subject correlation violates the mathematical assumption of independence. This artificially shrinks standard errors and drastically inflates Type I error rates (false positives). The classical solution is to use a mixed-effects model by adding a random intercept for each subject. 

However, in complex, non-linear Bayesian models—like the hierarchical piecewise regressions fitted by `smoothbp`—random effects introduce massive computational overhead. A model with 200 subjects requires 200 latent intercept parameters. The resulting posterior geometry can become highly "rugged," causing MCMC samplers to struggle with convergence, take hours to run, or require deep tuning of hyperparameters.

## Why use the `robustify()` approach?

The **Bayesian Sandwich** estimator implemented in `robustify()` provides a "best of both worlds" solution: the speed and stability of a marginal model, combined with the rigorous statistical validity of a mixed model.

Instead of explicitly modeling the random effects, you:
1. Fit a **marginal** `smoothbp()` model (which assumes independence and fits blazingly fast).
2. Call `robustify(fit, cluster = "Subject_ID")`.

Under the hood, `robustify()` calculates the frequentist robust sandwich covariance matrix (which penalizes for within-subject clustering). It then applies a multidimensional **affine transformation** to your MCMC chains. This mathematically "stretches" the posterior samples so their variance exactly matches the robust covariance. Downstream, your HDIs, plots, and Bayes Factors will automatically reflect the true, wider uncertainty.

### Assumptions and Limitations

While the Bayesian Sandwich estimator provides a powerful, computationally cheap alternative to heavy mixed-effects models, its defensibility hinges on several key mathematical assumptions:

**Valid and Defensible When:**
*   **Asymptotic Normality**: The affine transformation linearly scales and rotates the MCMC chains using covariance matrices (via Cholesky factorization). This assumes the unadjusted posterior distribution is roughly multivariate normal or elliptical.
*   **Large Number of Independent Clusters**: Just like frequentist GEEs (Generalized Estimating Equations), the empirical "meat" of the sandwich matrix relies on large-sample asymptotics across clusters. It works exceptionally well when you have a sufficient number of independent clusters (e.g., $N > 30$ subjects).
*   **Unbiased Marginal Means**: The sandwich estimator corrects the *variance* of the parameters, not the *point estimates* (posterior means). It assumes that omitting the random effects does not structurally bias the marginal mean trajectory.

**Not Valid or Defensible When:**
*   **Highly Skewed or Multimodal Posteriors**: If the underlying posterior geometry is strongly non-normal (e.g., severe "banana shapes", strict boundary constraints, or multimodality), applying a linear stretch can severely distort the tails or push MCMC draws into mathematically impossible parameter spaces (like negative variances).
*   **Small Number of Clusters**: If your dataset only contains a small handful of clusters (e.g., 4 hospital sites), the empirical score variance will be highly unstable, leading to a severely biased (often too narrow) robust covariance estimate.
*   **Strong Subject-Level Non-Linear Heterogeneity**: In highly non-linear models (like logistic/Poisson piecewise models with massive unmodeled random slopes), the marginal mean mathematically differs from the conditional mean. In these cases, fitting a naive marginal model yields structurally biased point estimates that no variance adjustment can fix. You must fit the full hierarchical model.

## Example: 2 Groups, 3 Time Points

Let's assume a basic clinical trial where the `Treatment` group experiences a shift in trajectory after Baseline.

```{r setup}
library(smoothbp)
library(dplyr)
library(ggplot2)
library(ggthemes) # For Tufte styling
# Ensure robustify is loaded if running interactively before install
if(!exists("robustify")) source("../R/robustify.R")

set.seed(42)
n_subjects <- 60
# Subject random intercepts (unmodeled by smoothbp, causing within-subject correlation)
subjects <- data.frame(
  Subject_ID = factor(1:n_subjects),
  Treatment = factor(rep(c("Control", "Treatment"), each = n_subjects/2)),
  u_i = rnorm(n_subjects, mean = 0, sd = 10)
)

clinical_data <- expand.grid(
  Subject_ID = subjects$Subject_ID, 
  time = c(0, 1, 2)
)
clinical_data <- merge(clinical_data, subjects, by = "Subject_ID")

# Generate trajectories: baseline score 50, slow decay of -2
clinical_data$score <- 50 + clinical_data$u_i - 2 * clinical_data$time

# Treatment group gets an accelerated decay (-15 slope) after time = 0.5
is_treated <- clinical_data$Treatment == "Treatment"
post_bp <- pmax(0, clinical_data$time - 0.5)
clinical_data$score[is_treated] <- clinical_data$score[is_treated] - 15 * post_bp[is_treated]

# Add measurement noise
clinical_data$score <- clinical_data$score + rnorm(nrow(clinical_data), 0, 3)

# 1. Fit the fast marginal model
fit_naive <- smoothbp(
  score ~ time,
  b1 = ~ 1,                        # Shared pre-breakpoint slope
  deltas = list(~ Treatment),      # Post-breakpoint slope change depends on Treatment
  omega = list(fixed(0.5)),      # Known discrete breakpoint exactly at time = 0.5
  rho = list(fixed(100)),        # Force sharp breakpoint (approximating infinite rho)
  data = clinical_data,
  iter = 2000, 
  warmup = 1000,
  chains = 2,
  .verbose = FALSE
)

# 2. Apply the Bayesian Sandwich
fit_robust <- robustify(fit_naive, cluster = "Subject_ID")

# 3. Fit the full Mixed-Effects Model (Random Intercepts)
fit_mixed <- smoothbp(
  score ~ time,
  b0 = ~ 1 + (1|Subject_ID),       # Add random intercept for baseline
  b1 = ~ 1,                        # Shared pre-breakpoint slope
  deltas = list(~ Treatment),      # Post-breakpoint slope change depends on Treatment
  omega = list(fixed(0.5)),      # Known discrete breakpoint exactly at time = 0.5
  rho = list(fixed(100)),        # Force sharp breakpoint (approximating infinite rho)
  data = clinical_data,
  iter = 2000, 
  warmup = 1000,
  chains = 2,
  .verbose = FALSE
)
```

### Hypothesis Testing with Bayes Factors

Because `robustify()` structurally transforms the underlying MCMC draws, the posterior density evaluates differently. You can pass the robustified object directly into the `hypothesis()` function to calculate the Savage-Dickey density ratio for a Bayes Factor (e.g., testing if the Treatment slope differs from zero).

```{r bf}
# Test if the Treatment alters the slope post-breakpoint
bf_robust <- hypothesis(fit_robust, "delta1_TreatmentTreatment < 0")
bf_mixed  <- hypothesis(fit_mixed, "delta1_TreatmentTreatment < 0")

cat("--- Robust Marginal Model ---\n")
print(bf_robust)

cat("\n--- Full Mixed-Effects Model ---\n")
print(bf_mixed)

# The Evidence Ratio (Bayes Factor) will be appropriately penalized 
# in the robust model, yielding similar inferential strength to the mixed model.
```

### Visualizing the Robust HDI (Tufte Style)

A core tenet of Edward Tufte's design philosophy is maximizing the data-ink ratio. We can visualize the robustified Highest Density Intervals (HDIs) cleanly without chartjunk using `ggthemes::theme_tufte()`.

```{r tufte_plot, fig.width=8, fig.height=5}
# Create a dummy population dataset (no Subject_ID) to predict average trajectories
newdata_pop <- expand.grid(Treatment = factor(c("Control", "Treatment")), time = seq(0, 2, by = 0.1))

# Extract predicted population trajectories with 95% HDIs for both models
preds_robust <- fitted(fit_robust, newdata = newdata_pop)
plot_data_robust <- bind_cols(newdata_pop, preds_robust) %>% mutate(Model = "Marginal (Robust)")

preds_mixed <- fitted(fit_mixed, newdata = newdata_pop)
plot_data_mixed <- bind_cols(newdata_pop, preds_mixed) %>% mutate(Model = "Mixed Effects")

plot_data <- bind_rows(plot_data_robust, plot_data_mixed)

ggplot(plot_data, aes(x = time, y = fitted_mean, color = Treatment, linetype = Model)) +
  # Tufte-style minimalist presentation
  geom_ribbon(aes(ymin = fitted_Q2.5, ymax = fitted_Q97.5, fill = Treatment), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.2) +
  theme_tufte(base_size = 14, ticks = FALSE) +
  scale_color_manual(values = c("Control" = "#7f8c8d", "Treatment" = "#c0392b")) +
  scale_fill_manual(values = c("Control" = "#7f8c8d", "Treatment" = "#c0392b")) +
  labs(
    x = "Time Point",
    y = "Clinical Score",
    title = "Comparison of Population Trajectories and HDIs",
    subtitle = "Robust Marginal vs. Mixed-Effects Model"
  ) +
  theme(legend.position = "right")
```
