---
title: "Variable Selection with Spike-and-Slab Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variable Selection with Spike-and-Slab Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## When to use spike-and-slab

The standard `smoothbp()` model assumes that every covariate in the `b2`
formula contributes to the slope change at the transition.  In practice you
may have several candidate covariates and want to learn *which* of them
genuinely modify the post-change slope.

`smoothbp_ss()` addresses this by placing a **spike-and-slab** prior on each
`b2` coefficient (Kuo & Mallick, 1998).  Each coefficient is associated with
a binary inclusion indicator $\gamma_k$:

- **$\gamma_k = 0$ (spike)**: the coefficient is exactly zero — no slope
  change attributable to covariate $k$.  The corresponding `omega` and `rho`
  coefficients (matched by name) are also zeroed, reflecting the structural
  constraint that if a covariate does not affect the slope change, it should
  not influence the change-point location or sharpness either.
- **$\gamma_k = 1$ (slab)**: the coefficient follows its normal slab
  prior and enters the mean function.

The posterior mean of each $\gamma_k$ is the **posterior inclusion
probability** (PIP), which quantifies the evidence that covariate $k$
participates in the slope change.

---

## Example 1: Identifying a null covariate

A minimal example with one true predictor and one noise covariate
illustrates the basic workflow.

### Simulate data

```{r sim-simple}
library(smoothbp)

set.seed(7314)

dat <- simulate_smoothbp(
  n_subj = 15, n_obs = 12,
  b0 = 5, b1 = -0.3, b2 = 0.8,
  omega = 3, rho = 4, sigma = 0.4, sigma_u = 0.5,
  tau_range = c(0, 6)
)

# x1 has a true b2 effect; x2 is pure noise
dat$x1 <- rnorm(nrow(dat), mean = dat$tau / 3)
dat$y  <- dat$y + 0.6 * dat$x1 *
  (dat$tau - 3) * plogis((dat$tau - 3) * 4)
dat$x2 <- rnorm(nrow(dat))
```

### Fit the spike-and-slab model

Pass `b2`, `omega`, and `rho` formulas that include both covariates.  The
`spike` argument controls the prior.  By default the intercept is always
included (never spiked); set `spike_intercept = TRUE` to test whether any
piecewise structure exists at all.

```{r fit-simple}
fit <- smoothbp_ss(
  formula = y ~ tau,
  b0    = ~ 1 + (1 | subject),
  b1    = ~ 1,
  b2    = ~ 1 + x1 + x2,
  omega = ~ 1 + x1 + x2,
  rho   = ~ 1,
  data  = dat,
  spike = prior_spike_slab(pi = 0.5, slab = prior_normal(0, 2)),
  priors = smoothbp_priors(omega = prior_normal(3, 2, lb = 0)),
  chains = 4, iter = 2000, warmup = 1000,
  cores = 4
)
```

### Inspect inclusion probabilities

```{r pip-simple}
pip(fit)
#> x1        x2
#> 0.997     0.142

print(fit)
```

A PIP near 1 for `x1` provides strong evidence of a genuine slope-change
effect; the low PIP for `x2` indicates the model can confidently exclude it.

### Diagnostics

The usual diagnostic tools work on spike-and-slab fits:

```{r diag-simple}
trace_plot(fit)
pp_check(fit)
```

Check that the `gamma_*` traces show healthy mixing (frequent switches
between 0 and 1).  If an indicator is stuck at 0 or 1 across all chains,
consider whether the slab is too wide relative to the effect size or whether
the data are simply uninformative for that covariate.

---

## Example 2: Learning the inclusion probability

When you have many covariates, the fixed prior inclusion probability $\pi$
can strongly influence the results.  Setting `learn_pi = TRUE` places a
Beta hyperprior on a shared $\pi$ and lets the data inform the overall
sparsity level.

### Simulate a sparse design

```{r sim-sparse}
set.seed(5218)

dat2 <- simulate_smoothbp(
  n_subj = 20, n_obs = 15,
  b0 = 5, b1 = -0.3, b2 = 0.6,
  omega = 3, rho = 4, sigma = 0.4, sigma_u = 0.5,
  tau_range = c(0, 6)
)

# 4 covariates: only x1 has a true b2 effect
dat2$x1 <- rnorm(nrow(dat2))
dat2$x2 <- rnorm(nrow(dat2))
dat2$x3 <- rnorm(nrow(dat2))
dat2$x4 <- rnorm(nrow(dat2))

d <- dat2$tau - 3
sig <- plogis(d * 4)
dat2$y <- dat2$y + 0.5 * dat2$x1 * d * sig
```

### Fit with a Beta hyperprior

A `Beta(1, 3)` prior on $\pi$ gently encourages sparsity — it has mean 0.25,
expressing a prior expectation that about one quarter of covariates
contribute.  The conjugate Beta-Bernoulli update

$$\pi \mid \gamma \sim \text{Beta}\!\left(a + \textstyle\sum_k \gamma_k,\; b + K - \textstyle\sum_k \gamma_k\right)$$

adapts $\pi$ each iteration, so the model can learn whether the data
support a dense or sparse configuration.

```{r fit-sparse}
fit2 <- smoothbp_ss(
  formula = y ~ tau,
  b0    = ~ 1 + (1 | subject),
  b1    = ~ 1,
  b2    = ~ 1 + x1 + x2 + x3 + x4,
  omega = ~ 1 + x1 + x2 + x3 + x4,
  rho   = ~ 1,
  data  = dat2,
  spike = prior_spike_slab(
    slab           = prior_normal(0, 1),
    learn_pi       = TRUE,
    a              = 1,
    b              = 3,
    spike_intercept = FALSE
  ),
  priors = smoothbp_priors(omega = prior_normal(3, 2, lb = 0)),
  chains = 4, iter = 3000, warmup = 1500,
  cores = 4
)
```

### Results

```{r pip-sparse}
pip(fit2)
#> x1        x2        x3        x4
#> 0.993     0.124     0.098     0.107

summary(fit2)
```

The learned $\pi$ posterior can be examined directly from the draws:

```{r pi-posterior}
library(posterior)

dm <- as_draws_matrix(fit2$draws)
pi_draws <- as.numeric(dm[, "pi"])

quantile(pi_draws, c(0.025, 0.5, 0.975))
#> 2.5%    50%  97.5%
#> 0.05   0.21   0.56
```

The posterior for $\pi$ has contracted toward sparsity, consistent with only
one of four covariates having a true effect.

---

## Practical guidance

### Choosing the slab width

The slab standard deviation controls the range of plausible non-zero effect
sizes.  A slab that is much wider than the posterior concentration of a
true effect will inflate PIPs for null covariates — a known property of the
Kuo & Mallick approach.  Match the slab width to your expected effect scale:

| Expected effects | Suggested slab |
|------------------|---------------|
| Small (< 0.5)   | `prior_normal(0, 0.5)` |
| Moderate (0.5–2) | `prior_normal(0, 1)` |
| Large / uncertain | `prior_normal(0, 2)` |

When in doubt, use `learn_pi = TRUE` with a sparsity-encouraging Beta
prior (e.g., `Beta(1, 3)`) to partially compensate for slab-width
sensitivity.

### Interpreting PIPs

There is no universal threshold, but the following scale is reasonable for
exploratory work (Barbieri & Berger, 2004):

| PIP | Interpretation |
|-----|---------------|
| > 0.95 | Strong evidence for inclusion |
| 0.75–0.95 | Moderate evidence |
| 0.25–0.75 | Inconclusive |
| < 0.25 | Little evidence for inclusion |

For formal variable selection, the **median probability model** includes all
covariates with PIP > 0.5.

### Structural zeroing

When $\gamma_k = 0$, the `omega` and `rho` coefficients sharing the same
name as the spiked `b2` coefficient are also set to zero.  This means you
can inspect the posterior of `omega` or `rho` covariates and expect exact
zeros in draws where the corresponding covariate is excluded — a useful
check that the model is behaving as intended.

---

## References

- Kuo, L. and Mallick, B. (1998). Variable selection for regression models.
  *Sankhya B*, 60(1):65--81.
- Barbieri, M. M. and Berger, J. O. (2004). Optimal predictive model
  selection. *Annals of Statistics*, 32(3):870--897.
